### 2017-09-16 caustic-assets/paper/main.tex

\documentclass[journal]{../styles/IEEEtran}

% Packages.
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}

\begin{document}

% Title.
\title{Caustic: Reinventing Database Transactions}
\author{Ashwin Madavan}
\maketitle

% Abstract.
\begin{abstract}
Database languages are designed for ease of implementation and not ease of use. For example, every relational database has its own interpretation of the SQL specification that is most convenient for the underlying database architecture and most often incompatible with its competition, and most key-value stores expose primtiive interfaces that are lacking in functionality and programmability. Caustic is a robust query language that is intuitive to use and straightforward to integrate with any database.
\end{abstract}

% Introduction
\section{Introduction}
The syntax of query languages varies wildly between databases. For example, relational databases like MySQL and PostgreSQL claim to implement the same SQL specification, but, in reality, all but the most trivial of queries are incompatible with each other. This lack of uniformity between languages tightly couples the database with the queries that are run on it, and makes it difficult to switch storage systems as requirements change.

The structure of query languages is deeply motivated by the capabilities and limits of the underlying database and not by what applications actually require. In particular, the language of relational algebra, which SQL is based upon, is not first-nature for most programmers. [On the Principles of Good Language Design, CJ Date] [http://wiki.c2.com/?SqlFlaws]

It's clear that query languages are typically designed for ease of implementation and not ease of use. They have disparate interfaces, deficient functionality, and unintuitive syntax. Caustic provides both a robust language for specifying transactions and an efficient runtime for executing them on arbitrary databases.

% Background
\section{Background}

% Runtime
\section{Runtime}
The runtime is responsible for executing transactions on arbitrary databases. Transactions are represented as abstract-syntax trees, because they are easy to manipulate, serialize, and optimize. The runtime exposes a Thrift interface for cross-language transaction execution, and is capable of running on any database that supports the following primitive operations:

\begin{itemize}
  \item \textbf{get}: Returns the values of a set of keys.
  \item \textbf{cput}: Conditionally apply changes iff dependent keys remain unchanged.
\end{itemize}

\subsection{Representation}

\subsection{Snapshot Isolation}

\subsection{Optimizations}
Because transactions are represented as abstract-syntax trees, the runtime is able to perform some interesting optimizations to improve execution performance. Most importantly, the runtime performs read and write batching. In other words, the runtime fetches \emph{as many reads as possible} and applies \emph{all writes together}. Clearly, the runtime performs optimally few roundtrips to and from the underlying key-value store, which is typically the performance bottleneck in database applications.

\subsection{Tunable Consistency}
Transactions are executed using MVCC, and derive their consistency and isolation guarantees from the underlying key-value store. If the underlying key-value has a repeatable isolation level, for example, then so will transactions in Caustic. If the underlying key-value is not transactional, then neither are transactions in Caustic. Therefore, applications may substitute a key-value store with weaker transactional guarantees (or no transactions at all!) to improve performance.

\subsection{Benchmarks}
Compare the Caustic runtime (running on RocksDB) to comparable systems like Calvin and Tango. Rely on quantitative metrics like end-to-end latency and transaction throughput on standard database benchmarks like YCSB and TPC-C.

\subsubsection{Effect of Transaction Size}
Performance scales linearly with transaction size. This is enabled by the tail recursive design of the runtime and can be verified empirically from benchmarks.

\subsubsection{Effect of Read or Write Skew}
Transactions with more reads than writes will have a different performance profiles than transactions with more writes than reads. However, this difference in performance is an artifact of the choice of underlying key-value store and not due to a read or write bias in the runtime. In other words, a read optimized key-value store will fare better with read-heavy transactions and a write optimized key-value store with write-heavy transactions in Caustic.

\subsubsection{Effect of Contention}
What happens to performance under contention? The number of retries required to execute a transaction may be modeled as a negative binomial distribution in which the probability of success $p$ is the contention probability. Therefore, $A = 1 + NB(1, p)$ is the distribution of attempts. Known results about the negative binomial distribution, may be used to make predictions about the mean and variance of transaction execution latency under contention. For example, the mean number of attempts $\bar{A} = 1 + \frac{p}{1 - p} = \frac{1}{1 - p}$. The contention probability is determined by a number of factors including: the number of keys that are read and written, the number of database reads and writes, and the latency of database reads and write.

% Syntax
\section{Syntax}
The Caustic syntax is a new high-level programming language that compiles into transactions on the Caustic runtime. Regardless of the features they might have (objects, functions, types, etc.), all programs eventually compile down into load, stores, and operations on memory - and memory is a key-value store. The purpose of Caustic syntax is to provide a rich language for expressing \emph{transactional} loads, stores, and operations on \emph{key-value stores}.

\subsection{Concurrency}
Because all programs written in Caustic compile into transactions, all programs are automatically thread-safe. No fancy synchronization mechanisms are required to build parallelizable programs in Caustic, everything is concurrent by default.

\subsection{Separation of Code and Data}
Every database has its own language for writing stored procedures. Even SQL databases typically support very different subsets of the SQL specification, that are not interoperable for anything but the most trivial of programs.  Because the syntax varies so wildly between different databases, stored procedures are very tightly coupled with the choice of underlying database. This makes it tremendously expensive to change databases once development is under way, and impossible to test stored procedures in isolation. Caustic provides a uniform interface over arbitrary databases. In other words, Caustic programs written for ione database will be identical to the same program written for any another. This allows programmers to select whichever database is most appropriate for their use-case (eg. in-memory for tests, MySQL for production).

\subsection{Interoperability}
Because of the simplicity of the Caustic grammar and runtime, the compiler is able to generate executable binaries that serve Caustic programs over a variety of mediums over any underlying database. Most OLTP applications can be divided into three parts: a database connector (eg. SQLAlchemy), a set of stored procedures (eg. Flask), and an external interface (eg. REST, Thrift). By specifying their stored procedures in Caustic, programmers can now rely on the Caustic compiler to generate the first and third parts for them.

\subsection{Benchmarks}
Compare the Caustic syntax qualitatively to SQL, Tango, and Java. Measure differences in code complexity (measured in Cyclomatic or Halstead Complexity), size of executables, and lines of code for a variety of sample use cases (eg. implementing ZooKeeper or BookKeeper).

\end{document}

### 2017-09-16 caustic-assets/review/review.tex

\documentclass[journal]{../styles/IEEEtran}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}

\begin{document}

% Title
\title{Caustic: Literature Review}
\author{Ashwin Madavan}
\date{15 September 2017}
\maketitle

% HyperDex
\section{HyperDex}

  \subsection{Hyperspace Hashing}
  Hyperspace Hashing colocates regions of the n-dimensional hyperspace induced by the attributes of 
  an object in order to perform efficient secondary indexing. Each column of a table corresponds to 
  a different dimension in a hyperspace. Each physical server is responsible for storing data within 
  a particular region in the hyperspace. By partitioning data in this manner, you can optimize 
  searches (which correspond to hyperplanes) by reducing the number of server that they need to 
  contact.

  \subsection{Value Dependent Chaining}
  Object updates occur in two linear passes over all servers containing an object. In the first 
  pass, each server creates a pending update and propagates the change to the next server. In the 
  second pass, each server commits the pending update. Linear object updates means that updates to 
  an object have to *sequentially* pass through every server that it is physically stored on. While 
  this commit protocol ensures consistency, it will not be as performant for larger objects.

  \subsection{Performance}
  I don't understand how the performance of their system is so significantly better than Cassandra. 
  I'm always skeptical of benchmarks, particularly when they show such significant improvements over 
  high-performance, enterprise-grade solutions. Something about these benchmarks just feels wrong. 
  Both HyperDex and Cassandra use a log-based storage engine (HyperDex stores a log file for each 
  object, and Cassandra uses LSM trees), so it probably isn't the on-disk representation that causes 
  the difference in performance.

  A closer look at the benchmarks reveals that Cassandra actually performs better on writes than 
  HyperDex (which is unsurprising given that HyperDex uses a linear commit protocol) and that the 
  higher read performance of HyperDex causes it benchmark better. Why might HyperDex have higher 
  read performance? In Cassandra, key-value pairs are replicated $f$ times, and, by default, 
  Cassandra returns the result of a read on one of the replicas of a key. In HyperDex, objects are 
  replicated $f$ times across $n$ subspaces. Therefore, there are $n$ times as many replicas of an 
  object as there are in Cassandra. Cassandra read throughput scales linearly with the number of 
  nodes. In the benchmark, objects in HyperDex are replicated to three subspaces (key subspace + two 
  attribute subspaces). If Cassandra had three times as many replicas, it would have three times the 
  read throughput. Clearly, the benchmarks seem biased toward HyperDex, because they are replicating 
  objects many more times than Cassandra.

% Yesquel
\section{Yesquel}

  \subsection{Decoupling Query Processor and Storage Engines}
  One of the main achievements of Yesquel is that it shows that dropping the SQL front end and 
  compromising consistency are not fundamentally necessary to achieve performance at scale. The
  paper acknowledges that,

    \begin{quote}
    Traditionally, the storage system had been a SQL database system, which is convenient for the
    developer. However, with the emergence of large Web applications in the past decade, the SQL
    database became a scalability bottleneck. Developers reacted in two ways. They used application
    specific techniques to improve scalability - which is a laborious and complicates application
    code. Or they replaced database systems with specialized NoSQL systems that scaled well by
    shedding functionality.
    \end{quote}

  However, by decoupling the query processor from the underlying storage engine, Yesquel was able to
  preserve a relational interface over a denormalized key-value store without sacrificing
  performance. I'm not convinced of the novelty of this, because it has been demonstrated before in
  a number of projects (Cassandra, CockroachDB, Aerospike, ArangoDB, etc.). Nonetheless, it is an
  important feature that deserves mention.

  \subsection{YDBT}
  Perhaps the most novel aspect of Yesquel is the Yesquel Distributed Balanced Tree (YDBT). My
  understanding is that it is functionally similar to a B+ tree, except that its nodes are split
  both when they grow too large and when they are under too high load.

  \subsection{Integrated Caching}
  This feature has already been implemented in Caustic, and it is reassuring that other people are
  doing the same. The idea is that transactions can cache incoherently without compromising data
  integrity, by speculating about the value of a key from a potentially stale version in cache and
  then validating these cached version on commit.

  \subsection{Performance}
  Benchmarks show that the system performs in the ballpark of Redis. While it is interesting that 
  their YDBTs perform similarly to hash-table systems, I'm still not convinced that they provide any
  additional functionality that cannot be directly encoded into a hash-table.

\end{document}

### 2017-09-23 caustic-runtime/src/main/scala/caustic/runtime/Database.scala
Migrated the LocalDatabase to Caffeine

@@ -1,10 +1,8 @@
 package caustic.runtime
 
 import Database._
-
 import org.apache.thrift.TException
 import org.apache.thrift.async.AsyncMethodCallback
-
 import scala.annotation.tailrec
 import scala.concurrent.{ExecutionContext, Future}
 import scala.util.{Failure, Success}

### 2017-09-23 caustic-runtime/src/main/scala/caustic/runtime/memory/LocalDatabase.scala

@@ -1,7 +1,9 @@
 package caustic.runtime
 package memory
 
-import scala.collection.mutable
+import com.github.benmanes.caffeine.{cache => caffeine}
+
+import scala.collection.JavaConverters._
 import scala.concurrent.{ExecutionContext, Future}
 
 /**
@@ -11,25 +13,29 @@ import scala.concurrent.{ExecutionContext, Future}
  *
  * @param underlying Underlying data store.
  */
-case class LocalDatabase(underlying: mutable.Map[Key, Revision]) extends Database {
+case class LocalDatabase(
+  underlying: caffeine.Cache[Key, Revision]
+) extends Database {
 
   override def get(keys: Set[Key])(
     implicit ec: ExecutionContext
   ): Future[Map[Key, Revision]] =
-    Future {
-      keys.map(k => k -> this.underlying.get(k))
-        .collect { case (k, Some(v)) => k -> v }
-        .toMap
-    }
+    Future(this.underlying.getAllPresent(keys.asJava).asScala.toMap)
 
   override def cput(depends: Map[Key, Version], changes: Map[Key, Revision])(
     implicit ec: ExecutionContext
   ): Future[Unit] =
     Future {
-      if (depends.exists { case (k, v) => this.underlying.get(k).exists(_.version > v) })
+      // Determine if the dependencies conflict with the underlying database.
+      val conflicts = this.underlying.getAllPresent(depends.keys.asJava).asScala.exists {
+        case (key, rev) => depends(key) < rev.version
+      }
+
+      // Throw an exception on conflict or perform updates otherwise.
+      if (conflicts)
         throw new Exception("Transaction conflicts.")
       else
-        this.underlying ++= changes
+        this.underlying.putAll(changes.asJava)
     }
 
 }
@@ -37,9 +43,10 @@ case class LocalDatabase(underlying: mutable.Map[Key, Revision]) extends Databas
 object LocalDatabase {
 
   /**
+   * Constructs an empty, in-memory database.
    *
-   * @return
+   * @return Empty LocalDatabase.
    */
-  def empty: LocalDatabase = LocalDatabase(mutable.Map.empty)
+  def empty: LocalDatabase = LocalDatabase(caffeine.Caffeine.newBuilder().build[Key, Revision]())
 
 }

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/Cache.scala
Updated caching
Mitigates the cache thrashing problem, by only invalidating the specific
keys that cause a conflict and not all keys that participate in the
transaction. There is still the vulnerability for thrashing, but at
least now it will only thrash on the keys that are actually under
contention. For very high contention situations, it might better to do
away with the cache entirely.

@@ -1,5 +1,8 @@
 package caustic.runtime
 
+import caustic.runtime.thrift.ConflictException
+
+import scala.collection.JavaConverters._
 import scala.concurrent.{ExecutionContext, Future}
 import scala.util.{Failure, Success}
 
@@ -72,9 +75,9 @@ trait Cache extends Database {
       case Success(_) =>
         // Update the values of changed keys.
         update(changes)
-      case Failure(e) =>
+      case Failure(e: ConflictException) =>
         // Invalidate cached keys to force them to reload from the database.
-        invalidate(depends.keySet union changes.keySet).transform(_ => Failure(e))
+        invalidate(e.keys.asScala.toSet).transform(_ => Failure(e))
     }
 
 }

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/Database.scala

@@ -1,12 +1,14 @@
 package caustic.runtime
 
 import Database._
-import org.apache.thrift.TException
+
 import org.apache.thrift.async.AsyncMethodCallback
+import org.apache.thrift.TException
+
 import scala.annotation.tailrec
+import scala.collection.mutable
 import scala.concurrent.{ExecutionContext, Future}
 import scala.util.{Failure, Success}
-import scala.collection.mutable
 
 /**
  * An asynchronous, transactional key-value store.
@@ -77,15 +79,12 @@ trait Database extends thrift.Database.AsyncIface {
 
       // Load the keys, update the local snapshot, and reduce the transaction. If the result is a
       // literal then return, otherwise recurse on the partially evaluated transaction.
-      get(keys) transformWith {
-        case Success(r) =>
-          snapshot ++= r
-          reduce(List(txn), List.empty) match {
-            case l: Literal => Future(l)
-            case o: Expression => evaluate(o)
-          }
-        case Failure(e) =>
-          Future.failed(new thrift.ReadException(e.getMessage))
+      get(keys) flatMap { r =>
+        snapshot ++= r
+        reduce(List(txn), List.empty) match {
+          case l: Literal => Future(l)
+          case o: Expression => evaluate(o)
+        }
       }
     }
 
@@ -173,10 +172,7 @@ trait Database extends thrift.Database.AsyncIface {
         buffer.clear()
         e.result
     } flatMap { r =>
-      cput(snapshot.mapValues(_.version).toMap, buffer.toMap).transformWith {
-        case Success(_) => Future(r)
-        case Failure(e) => Future.failed(new thrift.WriteException(e.getMessage))
-      }
+      cput(snapshot.mapValues(_.version).toMap, buffer.toMap).map(_ => r)
     }
   }
 

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/memory/LocalDatabase.scala

@@ -27,15 +27,15 @@ case class LocalDatabase(
   ): Future[Unit] =
     Future {
       // Determine if the dependencies conflict with the underlying database.
-      val conflicts = this.underlying.getAllPresent(depends.keys.asJava).asScala.exists {
-        case (key, rev) => depends(key) < rev.version
-      }
+      val conflicts = this.underlying
+        .getAllPresent(depends.keys.asJava).asScala
+        .filter { case (k, r) => depends(k) < r.version }
 
       // Throw an exception on conflict or perform updates otherwise.
-      if (conflicts)
-        throw new Exception("Transaction conflicts.")
-      else
+      if (conflicts.isEmpty)
         this.underlying.putAll(changes.asJava)
+      else
+        throw new thrift.ConflictException(conflicts.keySet.asJava)
     }
 
 }

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/relational/MySQLDatabase.scala

@@ -1,9 +1,8 @@
 package caustic.runtime
-package mysql
+package relational
 
 import java.sql.Connection
 import javax.sql.DataSource
-
 import scala.collection.mutable
 import scala.concurrent.ExecutionContext
 
@@ -18,7 +17,7 @@ case class MySQLDatabase(underlying: DataSource)(
 ) extends RelationalDatabase(underlying) {
 
   override def schema: String =
-    s""" CREATE TABLE IF NOT EXISTS `schema`(
+    s""" CREATE TABLE IF NOT EXISTS `caustic`(
        |   `key` varchar (200) NOT NULL,
        |   `version` BIGINT DEFAULT 0,
        |   `type` INT,
@@ -31,13 +30,13 @@ case class MySQLDatabase(underlying: DataSource)(
     // Load all the rows that match the keys.
     val statement = connection.prepareStatement(
       s""" SELECT `key`, `version`, `type`, `value`
-         | FROM `schema`
+         | FROM `caustic`
          | WHERE `key` IN (${List.fill(keys.size)("?").mkString(",")})
        """.stripMargin
     )
 
-    keys.zipWithIndex.foreach { case (key, index) =>
-      statement.setString(index + 1, key)
+    keys.zipWithIndex.foreach {
+      case (key, index) => statement.setString(index + 1, key)
     }
 
     // Parse the query result set into a buffer.
@@ -66,7 +65,7 @@ case class MySQLDatabase(underlying: DataSource)(
   def upsert(connection: Connection, key: Key, revision: Revision): Unit = {
     // Prepare the SQL statement.
     val statement = connection.prepareStatement(
-      s""" INSERT INTO `schema` (`key`, `version`, `type`, `value`)
+      s""" INSERT INTO `caustic` (`key`, `version`, `type`, `value`)
          | VALUES (?, ?, ?, ?)
          | ON DUPLICATE KEY UPDATE
          | `version` = VALUES(`version`),

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/relational/PostgreSQLDatabase.scala

@@ -1,9 +1,8 @@
 package caustic.runtime
-package postgres
+package relational
 
 import java.sql.Connection
 import javax.sql.DataSource
-
 import scala.collection.mutable
 import scala.concurrent.ExecutionContext
 
@@ -18,7 +17,7 @@ case class PostgreSQLDatabase(underlying: DataSource)(
 ) extends RelationalDatabase(underlying) {
 
   override def schema: String =
-    s""" CREATE TABLE IF NOT EXISTS schema(
+    s""" CREATE TABLE IF NOT EXISTS caustic (
        |   key varchar (1000) NOT NULL,
        |   version BIGINT DEFAULT 0,
        |   type INT,
@@ -31,13 +30,13 @@ case class PostgreSQLDatabase(underlying: DataSource)(
     // Load all the rows that match the keys.
     val statement = connection.prepareStatement(
       s""" SELECT key, version, type, value
-         | FROM schema
+         | FROM caustic
          | WHERE key IN (${List.fill(keys.size)("?").mkString(",")})
        """.stripMargin
     )
 
-    keys.zipWithIndex.foreach { case (key, index) =>
-      statement.setString(index + 1, key)
+    keys.zipWithIndex.foreach {
+      case (key, index) => statement.setString(index + 1, key)
     }
 
     // Parse the query result set into a buffer.
@@ -66,7 +65,7 @@ case class PostgreSQLDatabase(underlying: DataSource)(
   def upsert(connection: Connection, key: Key, revision: Revision): Unit = {
     // Prepare the SQL statement.
     val statement = connection.prepareStatement(
-      s""" INSERT INTO schema (key, version, type, value)
+      s""" INSERT INTO caustic (key, version, type, value)
          | VALUES (?, ?, ?, ?) ON CONFLICT (key)
          | DO UPDATE SET version = ?, type = ?, value = ?
      """.stripMargin

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/relational/RelationalDatabase.scala

@@ -1,7 +1,10 @@
 package caustic.runtime
+package relational
 
 import java.sql.Connection
 import javax.sql.DataSource
+
+import scala.collection.JavaConverters._
 import scala.concurrent.{ExecutionContext, Future, blocking}
 
 /**
@@ -90,16 +93,15 @@ abstract class RelationalDatabase(underlying: DataSource)(
     implicit ec: ExecutionContext
   ): Future[Unit] =
     transaction { con =>
-      // Determine whether or not the transaction conflicts.
+      // Determine if the dependencies conflict with the underlying database.
       val current = if (depends.isEmpty) Map.empty[Key, Revision] else select(con, depends.keySet)
-      val conflicts = depends.exists { case (k, v) => current.get(k).exists(_.version > v) }
+      val conflicts = depends.filter { case (k, v) => current.get(k).exists(_.version > v) }
 
-      if (!conflicts) {
-        // If the transaction does not conflict, then perform changes.
+      // Throw an exception on conflict or perform updates otherwise.
+      if (conflicts.isEmpty) {
         changes.foreach { case (k, r) => upsert(con, k, r) }
       } else {
-        // Otherwise, fail the put operation.
-        throw new Exception("Transaction conflicts.")
+        throw new thrift.ConflictException(conflicts.keySet.asJava)
       }
     }
 

### 2017-09-24 caustic-runtime/src/test/scala/caustic/runtime/relational/MySQLDatabaseTest.scala

@@ -1,7 +1,7 @@
 package caustic.runtime
-package mysql
+package relational
 
-import caustic.runtime.mysql.MySQLDatabase
+import caustic.runtime.relational.MySQLDatabase
 import com.mchange.v2.c3p0.ComboPooledDataSource
 import org.junit.runner.RunWith
 import org.scalatest.junit.JUnitRunner

### 2017-09-24 caustic-runtime/src/test/scala/caustic/runtime/relational/PostgresDatabaseTest.scala

@@ -1,4 +1,4 @@
-package caustic.runtime.postgres
+package caustic.runtime.relational
 
 import caustic.runtime.DatabaseTest
 import com.mchange.v2.c3p0.ComboPooledDataSource
@@ -23,7 +23,7 @@ class PostgresDatabaseTest extends DatabaseTest with BeforeAndAfterAll {
     // Delete all the table metadata.
     val con = this.pool.getConnection()
     val smt = con.createStatement()
-    smt.execute("DROP TABLE IF EXISTS schema")
+    smt.execute("DROP TABLE IF EXISTS caustic")
     con.close()
 
     // Run the tests.

### 2017-09-24 caustic-assets/paper/main.tex
Removed IEEE stylesheet
Syntax Update
Initial implementation of the Caustic compiler. Still deeply flawed, but
the general outline of it has been written.

@@
-\documentclass[journal]{../styles/IEEEtran}
+\documentclass[12pt]{article}
 
@@
 \begin{abstract}
-Database languages are designed for ease of implementation and not ease of use.
-For example, every relational database has its own interpretation of the SQL specification that is most convenient for the underlying database architecture and most often incompatible with its competition, and most key-value stores expose primtiive interfaces that are lacking in functionality and programmability.
-Caustic is a robust query language that is intuitive to use and straightforward to integrate with any database.
+Databases are either easy to use or to scale. For example, relational databases were popularized in
+large part because of the programmability of SQL. However, this programmability comes at an
+incredible expense. Relational databases are notoriously difficult to support at scale, and so
+developers have increasingly turned toward more specialized NoSQL systems that scale well be
+shedding functionality. Developers are not only forced to choose between productivity and
+performance, but also stark differences between query languages makes their choice of database
+effectively permanent. Even databases that claim to support the same SQL standard only implement
+incompatible subsets of its functionality. The lack of a truly uniform interface tightly couples
+databases and the procedures that are executed against them. Caustic is both a syntax for expressing
+transactions and a runtime for executing them on arbitrary key-value stores, that is both
+straightforward to use and simple to integrate.
 \end{abstract}
-
-% Introduction
-\section{Introduction}
-The syntax of query languages varies wildly between databases.
-For example, relational databases like MySQL and PostgreSQL claim to implement the same SQL specification, but, in reality, all but the most trivial of queries are incompatible with each other.
-This lack of uniformity between languages tightly couples the database with the queries that are run on it, and makes it difficult to switch storage systems as requirements change.
-
-The structure of query languages is deeply motivated by the capabilities and limits of the underlying database and not by what applications actually require.
-In particular, the language of relational algebra, which SQL is based upon, is not first-nature for most programmers.
-[On the Principles of Good Language Design, CJ Date] [http://wiki.c2.com/?SqlFlaws]
-
-It's clear that query languages are typically designed for ease of implementation and not ease of use.
-They have disparate interfaces, deficient functionality, and unintuitive syntax.
-Caustic provides both a robust language for specifying transactions and an efficient runtime for executing them on arbitrary databases.
-
-% Background
-\section{Background}
 
@@
 \section{Runtime}
-The runtime is responsible for executing transactions on arbitrary databases.
-Transactions are represented as abstract-syntax trees, because they are easy to manipulate, serialize, and optimize.
-The runtime exposes a Thrift interface for cross-language transaction execution, and is capable of running on any database that supports the following primitive operations:
+  The runtime is responsible for executing transactions on arbitrary key-value stores. Runtimes
+  implement Multiversion Concurrency Controls over any key-value store that satisfies an minimal
+  interface.
+
+  \subsection{Representation}
+    A \texttt{Transaction} is represented by an abstract-syntax tree. A \texttt{Literal} corresponds
+    to a leaf of this syntax tree, and an \texttt{Expression} corresponds to a node. The runtime
+    respects three different types of literals (\texttt{Real}, \texttt{Text}, and \texttt{Flag}) and
+    thirty different kinds of expressions. With just these types and operations, the runtime is able
+    to emulate a modern programming language complete with conditional branching, loops, and local
+    variables.
+
+  \subsection{Execution}
+    Transactions are executed through repeated partial evaluation, according to the following
+    procedure.
 
     \begin{itemize}
-  \item \textbf{get}: Returns the values of a set of keys.
-  \item \textbf{cput}: Conditionally apply changes iff dependent keys remain unchanged.
+      \item \textbb{Fetch}: Call \texttt{get} on all keys that are read and written by the
+        transaction and add the returned revisions to a local snapshot. In order to guarantee
+        snapshot isolation, \texttt{get} is only called when a key is read or written for the
+        first time. By batching reads and writes and avoiding duplicate fetches, the runtime is
+        guaranteed to perform a minimal number of roundtrips to and from the database.
+      \item \textbb{Evaluate}: Recursively replace all expressions that have literal operands with
+        their corresponding literal result. For example, add(real(1), sub(real(0), real(2)))
+        returns real(-1). The result of all write expressions is saved to a local buffer and all
+        read expressions return the latest value of the key in either the local buffer or snapshot.
+      \item \textbb{Repeat}: Re-evaluate the transaction until it reduces to a single literal
+        value. Because all expressions with literal operands return a literal value, all
+        transactions eventually reduce to a literal.
     \end{itemize}
 
-\subsection{Representation}
+  \subsection{Integrated Caching}
+    MVCC allows the runtime to cache incoherently without sacrificing data integrity. The runtime
+    speculates about the value of a key by reading a potentially stale version from cache, and then
+    validate the cached versions when they commit. Multilevel, write-through caching is directly
+    integrated into the runtime:
 
-\subsection{Snapshot Isolation}
+    \begin{itemize}
+      \item Keys are updated on cache miss.
+      \item Keys are automatically updated after successful transactions.
+      \item Keys are automatically invalidated after conflicting transactions.
+    \end{itemize}
 
-\subsection{Optimizations}
-Because transactions are represented as abstract-syntax trees, the runtime is able to perform some interesting optimizations to improve execution performance.
-Most importantly, the runtime performs read and write batching.
-In other words, the runtime fetches \emph{as many reads as possible} and applies \emph{all writes together}.
-Clearly, the runtime performs optimally few roundtrips to and from the underlying key-value store, which is typically the performance bottleneck in database applications.
+    \subsubsection{Effect of Contention}
+      While the implementation of caching is fault tolerant, it is particularly vulnerable to
+      thrashing under contention. Consider if runtime A and runtime B read a key K from cache. A
+      updates the value of K, and then subsequently updates the value of K in cache. When B tries to
+      update the value of K it will cause a conflict, and so it will subsequently invalidate K in
+      cache.
 
-\subsection{Tunable Consistency}
-Transactions are executed using MVCC, and derive their consistency and isolation guarantees from the underlying key-value store.
-If the underlying key-value has a repeatable isolation level, for example, then so will transactions in Caustic.
-If the underlying key-value is not transactional, then neither are transactions in Caustic.
-Therefore, applications may substitute a key-value store with weaker transactional guarantees (or no transactions at all!) to improve performance.
+      A potential solution might be to remove the invalidation step from the procedure; however, the
+      implementation would no longer be fault tolerant. Consider if A updates K but fails to write
+      the new value to cache. Now the value in cache will remain stale until its evicted, but a key
+      under high contention may never be evicted from cache.
 
-\subsection{Benchmarks}
-Compare the Caustic runtime (running on RocksDB) to comparable systems like Calvin and Tango.
-Rely on quantitative metrics like end-to-end latency and transaction throughput on standard database benchmarks like YCSB and TPC-C.
+  \subsection{Performance}
 
     \subsubsection{Effect of Transaction Size}
-Performance scales linearly with transaction size. This is enabled by the tail recursive design of the runtime and can be verified empirically from benchmarks.
+      The runtime has been empirically shown to scale linearly with transaction size.
 
     \subsubsection{Effect of Read or Write Skew}
-Transactions with more reads than writes will have a different performance profiles than transactions with more writes than reads.
-However, this difference in performance is an artifact of the choice of underlying key-value store and not due to a read or write bias in the runtime.
-In other words, a read optimized key-value store will fare better with read-heavy transactions and a write optimized key-value store with write-heavy transactions in Caustic.
+      The quantity of read and write expressions will surely impact the performance profile of the
+      runtime. However, the difference in performance is an artifact of the choice of underlying
+      key-value store and not due to any inherent bias in the runtime.
 
     \subsubsection{Effect of Contention}
-What happens to performance under contention?
-The number of retries required to execute a transaction may be modeled as a negative binomial distribution in which the probability of success $p$ is the contention probability.
-Therefore, $A = 1 + NB(1, p)$ is the distribution of attempts.
-Known results about the negative binomial distribution, may be used to make predictions about the mean and variance of transaction execution latency under contention.
-For example, the mean number of attempts $\bar{A} = 1 + \frac{p}{1 - p} = \frac{1}{1 - p}$.
-The contention probability is determined by a number of factors including: the number of keys that are read and written, the number of database reads and writes, and the latency of database reads and write.
+      What happens to performance under contention? The number of retries required to execute a
+      transaction may be modeled as a negative binomial distribution in which the probability of
+      success $p$ is the contention probability. Therefore, $A = 1 + NB(1, p)$ is the distribution
+      of attempts. Known results about the negative binomial distribution, may be used to make
+      predictions about the mean and variance of transaction execution latency under contention. For
+      example, the mean number of attempts $\bar{A} = 1 + \frac{p}{1 - p} = \frac{1}{1 - p}$. The
+      contention probability is determined by a number of factors including: the number of keys that
+      are read and written, the number of database reads and writes, and the latency of database
+      reads and write.
 
@@
 \section{Syntax}
-The Caustic syntax is a new high-level programming language that compiles into transactions on the Caustic runtime.
-Regardless of the features they might have (objects, functions, types, etc.), all programs eventually compile down into load, stores, and operations on memory - and memory is a key-value store.
-The purpose of Caustic syntax is to provide a rich language for expressing \emph{transactional} loads, stores, and operations on \emph{key-value stores}.
+  While the language exposed by the runtime is robust and fully-featured, it can become verbose and
+  unintuitive as transactions get larger in size and broader in scope. In this section, we'll
+  introduce the Caustic programming language.
 
-\subsection{Concurrency}
-Because all programs written in Caustic compile into transactions, all programs are automatically thread-safe.
-No fancy synchronization mechanisms are required to build parallelizable programs in Caustic, everything is concurrent by default.
+  \subsection{Why Caustic?}
+    Relational databases have remained dominant particularly because of the programmability of SQL.
+    SQL provides developers with robust schemas that make it easy to structure and debug database
+    queries. However, relational database are notoriously difficult to support at scale, and so
+    developers are increasingly turning towards more specialized NoSQL systems that scale well by
+    shedding functionality.
 
-\subsection{Separation of Code and Data}
-Every database has its own language for writing stored procedures.
-Even SQL databases typically support very different subsets of the SQL specification, that are not interoperable for anything but the most trivial of programs.
-Because the syntax varies so wildly between different databases, stored procedures are very tightly coupled with the choice of underlying database.
-This makes it tremendously expensive to change databases once development is under way, and impossible to test stored procedures in isolation.
-Caustic provides a uniform interface over arbitrary databases.
-In other words, Caustic programs written for ione database will be identical to the same program written for any another.
-This allows programmers to select whichever database is most appropriate for their use-case (eg.
-in-memory for tests, MySQL for production).
+    While SQL has previously enjoyed remarkable popularity, the language is deeply flawed. In his
+    letter entitled "Some Principles of Good Language Design", CJ Date outlines a number of these
+    flaws as well as a criteria for evaluating languages in general. Date argues that well-designed
+    languages should have a consistent syntax, a canonical implementation, a terse grammar, and an
+    intuitive structure and provides examples that show how SQL fails to satisfy each of these
+    fundamental requirements.
+
+  \subsection{Automatic Concurrency}
+    Because all programs written in Caustic compile into transactions on the Caustic runtime, all
+    programs are automatically thread-safe. Developers no longer have to worry about fancy
+    synchronization mechanisms in order to build parallelizable programs.
+
+  \subsection{Uniform Interface}
+    Every database has its own query language. Even databases that implement the same SQL
+    specification usually support disparate subsets of the language. This has the effect of tightly
+    coupling the underlying database and the procedures that are run against it. Caustic provides
+    a uniform interface over arbitrary databases, which allows programmers to select whichever
+    underlying database is most appropriate for their use-case without modifying the code in any
+    way shape or form.
 
   \subsection{Interoperability}
-Because of the simplicity of the Caustic grammar and runtime, the compiler is able to generate executable binaries that serve Caustic programs over a variety of mediums over any underlying database.
-Most OLTP applications can be divided into three parts: a database connector (eg.
-SQLAlchemy), a set of stored procedures (eg.
-Flask), and an external interface (eg.
-REST, Thrift).
-By specifying their stored procedures in Caustic, programmers can now rely on the Caustic compiler to generate the first and third parts for them.
-
-\subsection{Benchmarks}
-Compare the Caustic syntax qualitatively to SQL, Tango, and Java.
-Measure differences in code complexity (measured in Cyclomatic or Halstead Complexity), size of executables, and lines of code for a variety of sample use cases (eg.
-implementing ZooKeeper or BookKeeper).
+    Because of the simplicity of the Caustic grammar and runtime, the compiler is able to generate
+    executable binaries that serve Caustic programs over a variety of mediums over any underlying
+    database. Most OLTP applications can be divided into three parts: a database connector
+    (eg. SQLAlchemy), a set of stored procedures (eg. Flask), and an external interface
+    (eg. REST, Thrift). By specifying their stored procedures in Caustic, programmers can now rely
+    on the Caustic compiler to generate the first and third parts for them.
 

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/memory/LocalCache.scala

@@ -2,7 +2,9 @@ package caustic.runtime
 package memory
 
 import com.github.benmanes.caffeine.{cache => caffeine}
+
 import java.util.concurrent.TimeUnit
+
 import scala.collection.JavaConverters._
 import scala.concurrent.duration.Duration
 import scala.concurrent.{ExecutionContext, Future}

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/relational/PostgreSQLDatabase.scala

@@ -3,6 +3,7 @@ package relational
 
 import java.sql.Connection
 import javax.sql.DataSource
+
 import scala.collection.mutable
 import scala.concurrent.ExecutionContext
 

### 2017-09-24 caustic-syntax/src/main/python/caustic/syntax/codec.py

@@ -1,115 +1,0 @@
-import math
-
-from caustic.runtime.thrift.ttypes import *
-
-
-# Record kinds.
-Structure = 'structure'
-Attribute = 'attribute'
-Reference = 'reference'
-
-
-# Reserved Delimiters.
-FieldDelimiter = '@'
-ArrayDelimiter = ','
-
-
-# Implicit Conversions.
-def fmt(x):
-    if x is None:
-        return Empty
-    elif isinstance(x, Transaction):
-        return x
-    elif isinstance(x, str):
-        return text(x)
-    elif isinstance(x, int):
-        return real(x)
-    elif isinstance(x, float):
-        return real(x)
-    elif isinstance(x, bool):
-        return flag(x)
-    else:
-        raise TypeError('Unknown type for {}'.format(x))
-
-
-# Literal Values.
-def flag(x): return Transaction(literal=Literal(flag=x))
-def real(x): return Transaction(literal=Literal(real=x))
-def text(x): return Transaction(literal=Literal(text=x))
-
-
-# Core Expressions.
-def read(k): return Transaction(expression=Expression(read=Read(fmt(k))))
-def write(k, v): return Transaction(expression=Expression(write=Write(fmt(k), fmt(v))))
-def load(k): return Transaction(expression=Expression(load=Load(fmt(k))))
-def store(n, v): return Transaction(expression=Expression(store=Store(fmt(n), fmt(v))))
-def branch(c, p, f=None): return Transaction(expression=Expression(branch=Branch(fmt(c), fmt(p), fmt(f))))
-def cons(a, b): return Transaction(expression=Expression(cons=Cons(fmt(a), fmt(b))))
-def prefetch(k): return Transaction(expression=Expression(prefetch=Prefetch(fmt(k))))
-def repeat(c, b): return Transaction(expression=Expression(repeat=Repeat(fmt(c), fmt(b))))
-def rollback(m): return Transaction(expression=Expression(rollback=Rollback(fmt(m))))
-def block(x, *rest): return reduce(lambda a, b: cons(a, b), rest, x)
-
-
-# String Expressions.
-Empty = text("")
-
-def contains(x, y): return Transaction(expression=Expression(contains=Contains(fmt(x), fmt(y))))
-def length(x): return Transaction(expression=Expression(length=Length(fmt(x))))
[…]

### 2017-09-24 caustic-syntax/src/main/python/caustic/syntax/operations.py

@@ -1,54 +1,0 @@
-import abc
-
-from caustic.syntax.codec import *
-
-
-class InfixOperations(object):
-    """
-
-    """
-    __metaclass__ = abc.ABCMeta
-
-    # String Operations.
-    def __len__(self): return length(self)
-    def __contains__(self, y): return contains(self, text(y))
-    def matches(self, y): return matches(self, text(y))
-    def index(self, y): return indexOf(self, text(y))
-    def __getitem__(self, item): pass
-
-    # Math Operations.
-    def __neg__(self): return sub(Zero, self)
-    def __abs__(self): return abs(self)
-    def __ceil__(self): return ceil(self)
-    def __floor__(self): return floor(self)
-    def __trunc__(self): return trunc(self)
-    def __round__(self): return round(self)
-    def __add__(self, y): return add(self, y)
-    def __radd__(self, x): return add(x, self)
-    def __sub__(self, y): return sub(self, y)
-    def __rsub__(self, x): return sub(x, self)
-    def __mul__(self, y): return mul(self, y)
-    def __rmul__(self, x): return mul(x, self)
-    def __truediv__(self, y): return div(self, y)
-    def __rtruediv__(self, x): return div(x, self)
-    def __floordiv__(self, y): return floor(div(self, y))
-    def __rfloordiv__(self, x): return floor(div(x, self))
-    def __mod__(self, y): return mod(self, y)
-    def __rmod__(self, x): return mod(x, self)
-    def __pow__(self, y, mod=None):  return mod(pow(self, y), mod) if mod else pow(self, y)
-    def __rpow__(self, x):  return pow(x, self)
-
-    # Comparison Operations.
-    def __invert__(self): return negate(self)
-    def __eq__(self, y): return eq(self, y)
-    def __ne__(self, y): return ne(self, y)
-    def __lt__(self, y): return lt(self, y)
-    def __le__(self, y): return le(self, y)
-    def __gt__(self, y): return gt(self, y)
-    def __ge__(self, y): return ge(self, y)
-    def __and__(self, y): return both(self, flag(y))
-    def __rand__(self, x): return both(x, self)
-    def __xor__(self, y): return xor(self, y)
-    def __rxor__(self, x): return xor(x, self)
-    def __or__(self, y): return either(self, y)
-    def __ror__(self, x): return either(x, self)

### 2017-09-24 caustic-syntax/src/main/python/caustic/syntax/record.py

@@ -1,123 +1,0 @@
-from __future__ import division
-
-from caustic.syntax.codec import *
-
-
-class Record(object):
-    """
-
-    """
-
-    def __init__(self, key):
-        self.key = key
-        self.kind = Record(key + FieldDelimiter + 'kind')
-
-    def __getattr__(self, name):
-        return Record(branch(
-            self.kind == self.Reference,
-            read(self.key) + FieldDelimiter + name,
-            self.key + FieldDelimiter + name
-        ))
-
-    def __setattr__(self, name, value):
-        pass
-
-    def __del__(self):
-        pass
-
-    def __delattr__(self, name):
-        pass
-
-    # String Operations.
-    def __len__(self):
-        # len(x) -> length(x)
-        return length(read(self.key))
-
-    def __contains__(self, y):
-        # 'foo' in x -> contains(x, 'foo')
-        return contains(read(self.key), text(y))
-
-    def matches(self, y):
-        # x.matches('[a-z]+') -> matches(x, '[a-z]+')
-        return matches(read(self.key), text(y))
-
-    def index(self, y):
-        # x.index('foo') -> indexOf(x, 'foo')
-        return indexOf(read(self.key), text(y))
-
-    def __getitem__(self, y):
-        # x[1:4] -> slice(x, 1, 4)
-        if isinstance(y, slice):
-            return slice(read(self.key), y.start if y.start else 0, y.stop if y.stop else len(self))
-        else:
-            return slice(read(self.key), y, y + 1)
-
-    # Math Operations.
-    def __neg__(self):
-        return sub(Zero, read(self.key))
-    def __abs__(self):
-        return abs(read(self.key))
-    def __ceil__(self):
-        return ceil(read(self.key))
-    def __floor__(self):
-        return floor(read(self.key))
-    def __trunc__(self):
-        return branch(less(read(self.key), Zero), ceil(read(self.key)), floor(read(self.key)))
-    def __round__(self):
-        return round(read(self.key))
-    def __add__(self, y):
-        return add(read(self.key), real(y))
-    def __radd__(self, x):
-        return add(real(x), read(self.key))
-    def __sub__(self, y):
-        return sub(read(self.key), real(y))
-    def __rsub__(self, x):
-        return sub(real(x), read(self.key))
-    def __mul__(self, y):
-        return mul(read(self.key), real(y))
-    def __rmul__(self, x):
-        return mul(real(x), read(self.key))
-    def __truediv__(self, y):
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/Codec.scala

@@ -3,7 +3,9 @@ package caustic.syntax
 import caustic.runtime.thrift
 
 /**
- *
+ * A static library for generating Thrift transactions. Used by the generated sources to dynamically
+ * build and execute transactions on the underlying database. Must be imported into all source files
+ * in order for them to compile.
  */
 object Codec {
 
@@ -69,7 +71,7 @@ object Codec {
     thrift.Transaction.expression(thrift.Expression.floor(new thrift.Floor(x)))
 
   def abs(x: thrift.Transaction): thrift.Transaction =
-    branch(less(x, Zero), sub(Zero, x), x)
+    branch(lt(x, Zero), sub(Zero, x), x)
   def exp(x: thrift.Transaction): thrift.Transaction =
     pow(E, x)
   def tan(x: thrift.Transaction): thrift.Transaction =
@@ -95,9 +97,9 @@ object Codec {
   def sqrt(x: thrift.Transaction): thrift.Transaction =
     pow(x, Half)
   def ceil(x: thrift.Transaction): thrift.Transaction =
-    branch(equal(x, floor(x)), x, add(floor(x), One))
+    branch(eq(x, floor(x)), x, add(floor(x), One))
   def round(x: thrift.Transaction): thrift.Transaction =
-    branch(less(sub(x, floor(x)), Half), floor(x), ceil(x))
+    branch(lt(sub(x, floor(x)), Half), floor(x), ceil(x))
 
   // String Expressions.
   def contains(x: thrift.Transaction, y: thrift.Transaction): thrift.Transaction =
@@ -114,24 +116,23 @@ object Codec {
     thrift.Transaction.expression(thrift.Expression.indexOf(new thrift.IndexOf(x, y)))
 
   // Logical Expressions.
-  def both(x: thrift.Transaction, y: thrift.Transaction): thrift.Transaction =
+  def and(x: thrift.Transaction, y: thrift.Transaction): thrift.Transaction =
     thrift.Transaction.expression(thrift.Expression.both(new thrift.Both(x, y)))
-  def either(x: thrift.Transaction, y: thrift.Transaction): thrift.Transaction =
+  def or(x: thrift.Transaction, y: thrift.Transaction): thrift.Transaction =
     thrift.Transaction.expression(thrift.Expression.either(new thrift.Either(x, y)))
-  def negate(x: thrift.Transaction): thrift.Transaction =
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/Compiler.scala

@@ -13,13 +13,13 @@ trait Compiler extends CausticBaseVisitor[String] {
   /**
    *
    * @param caustic Caustic source code.
-   * @return
+   * @return Generated code.
    */
   def compile(caustic: CharStream): String = {
     val lexer = new CausticLexer(caustic)
     val tokens = new CommonTokenStream(lexer)
     val parser = new CausticParser(tokens)
-    this.visitService(parser.service())
+    this.visitProgram(parser.program())
   }
 
   final def compile(code: String): String =
@@ -30,5 +30,6 @@ trait Compiler extends CausticBaseVisitor[String] {
 
   final def compile(source: Source): String =
     try compile(source.mkString) finally source.close()
+
 
 }

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/Language.scala

@@ -1,124 +1,0 @@
-package caustic.syntax
-
-import Language._
-
-trait Language {
-
-  /**
-   * Conditionally branches on the result of the specified condition. An else clause may be
-   * optionally specified. Utilizes structural types, which can be enabled by importing the
-   * scala.language.reflectiveCalls language feature.
-   *
-   * @param condition Condition to branch on.
-   * @param success Execute if condition is satisfied.
-   * @param ctx Implicit transaction context.
-   * @return Optional else clause.
-   */
-  def If(condition: Transaction)(success: => Unit)(
-    implicit ctx: Context
-  ) = new {
-    private val before = ctx.txn
-    ctx.txn = Empty
-    success
-    private val pass = ctx.txn
-    ctx.txn = before
-    ctx += branch(condition, pass, Empty)
-
-    def Else(failure: => Unit): Unit = {
-      ctx.txn = Empty
-      failure
-      val fail = ctx.txn
-      ctx.txn = before
-      ctx += branch(condition, pass, fail)
-    }
-  }
-
-  /**
-   * Repeatedly performs the specified block while the condition is satisfied.
-   *
-   * @param condition Condition to loop on.
-   * @param block Loop body.
-   * @param ctx Implicit transaction context.
-   */
-  def While(condition: Transaction)(block: => Unit)(
-    implicit ctx: Context
-  ): Unit = {
-    val before = ctx.txn
-    ctx.txn = Empty
-    block
-    val body = ctx.txn
-    ctx.txn = before
-    ctx += repeat(condition, body)
-  }
-
-  /**
-   * Repeatedly performs the specified block over the interval, storing the value of the loop index
-   * in the provided variable.
-   *
-   * @param variable Loop variable.
-   * @param in Loop interval.
-   * @param block Loop body.
-   * @param ctx Implicit transaction context.
-   */
-  def For(variable: Transaction, in: Interval)(block: => Unit)(
-    implicit ctx: Context
-  ): Unit = {
-    ctx.set(variable, in.start)
-    val condition = in match {
-      case Interval(_, end, _, true)  => load(variable) <= end
-      case Interval(_, end, _, false) => load(variable) <  end
-    }
-
-    val before = ctx.txn
-    ctx.txn = Empty
-    block
-    ctx.set(variable, load(variable) + in.step)
-    val body = ctx.txn
-    ctx.txn = before
-    ctx += repeat(condition, body)
-  }
-
-  /**
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/ThriftCompiler.scala

@@ -1,0 +1,75 @@
+package caustic.syntax
+
+import caustic.grammar.CausticParser._
+import scala.collection.JavaConverters._
+
+/**
+ * A compiler from Caustic programs into Thrift IDLs.
+ */
+class ThriftCompiler extends Compiler {
+
+  override def visitProgram(ctx: ProgramContext): String = {
+    val imports = ctx.module().asScala.drop(1).map(x => s"""import "${x.getText}"""").mkString("\n")
+    val declarations = ctx.declaration().asScala.map(visitDeclaration)
+    val structs = declarations.filter(_ startsWith "struct").mkString("\n")
+    val services = declarations.filter(_ startsWith "service").mkString("\n")
+
+    // Serialize Caustic programs as a Thrift IDL.
+    s"""namespace * ${ctx.module(0).getText}
+       |$imports
+       |$structs
+       |$services
+     """.stripMargin
+  }
+
+  override def visitRecord(ctx: RecordContext): String = {
+    val name = ctx.Identifier(0).getText
+    val params = visitParameters(ctx.parameters())
+
+    // Serialize records as Thrift structs.
+    s"""struct $name {
+       |  $params
+       |}
+     """.stripMargin
+  }
+
+  override def visitService(ctx: ServiceContext): String = {
+    val name = ctx.Identifier(0).getText
+    val functions = ctx.function().asScala.map(visitFunction)
+
+    // Serialize services as Thrift services.
+    s"""service $name {
+       |$functions
+       |}
+     """.stripMargin
+  }
+
+  override def visitFunction(ctx: FunctionContext): String = {
+    val name = ctx.Identifier().getText
+    val args = visitParameters(ctx.parameters())
+    val ret  = visitType(ctx.`type`())
+
+    // Serialize functions as Thrift functions.
+    s"""$ret $name(
+       |  $args
+       |)
+     """.stripMargin
+  }
+
+  override def visitParameters(ctx: ParametersContext): String =
+    ctx.parameter().asScala
+      .zipWithIndex
+      .map { case (p, i) => s"${i + 1}: ${visitType(p.`type`())} ${p.Identifier()}," }
+      .mkString("\n")
+
+  override def visitType(ctx: TypeContext): String =
+    ctx.Identifier().getText match {
+      case _ if ctx.Ampersand() != null => "string"
+      case "Boolean" => "bool"
+      case "Integer" => "i64"
+      case "Decimal" => "double"
+      case "String"  => "string"
+      case _ => _
+    }
+
+}

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/http/HttpCompiler.scala

@@ -1,17 +1,0 @@
-package caustic.syntax
-package http
-
-/**
- * A compiler that generates a restful HTTP interface. For some foo.acid, the HttpCompiler generates
- * an executable HttpServer that serves a route for each function. Functions with only primitive or
- * reference arguments are generated as GET methods and all others are generated as POST methods.
- * The HttpCompiler gives Caustic out-of-box interoperability with browsers and shells.
- */
-class HttpCompiler extends Compiler {
-
-}
-
-object HttpCompiler {
-
-
-}

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/package.scala

@@ -1,95 +1,0 @@
-package caustic
-
-import caustic.runtime.thrift
-import caustic.runtime.thrift._
-import caustic.syntax.Language._
-import caustic.syntax.ops._
-
-import scala.language.implicitConversions
-
-package object syntax extends Language {
-
-  type Transaction = thrift.Transaction
-
-  // Codec.
-  implicit def lit2txn(x: Literal): Transaction = thrift.Transaction.literal(x)
-  implicit def exp2txn(x: Expression): Transaction = thrift.Transaction.expression(x)
-
-  // Language.
-  implicit def txn2ops(x: Transaction): TransactionOps = TransactionOps(x)
-  implicit def rec2txn(x: Record): Transaction = read(x.key)
-  implicit def rec2top(x: Record): TransactionOps = TransactionOps(x)
-  implicit def rec2ops(x: Record): RecordOps = RecordOps(x)
-  implicit def rng2int(x: Range): Interval = Interval(x.start, x.end, x.step, x.isInclusive)
-
-  // Literal Values.
-  def flag(x: Boolean): Transaction = Literal.flag(x)
-  implicit def boolean2flag(x: Boolean): Transaction = flag(x)
-  def real[T](x: T)(implicit num: Numeric[T]): Transaction = Literal.real(num.toDouble(x))
-  implicit def numeric2real[T](x: T)(implicit num: Numeric[T]): Transaction = real(x)
-  def text(x: String): Transaction = Literal.text(x)
-  implicit def string2text(x: String): Transaction = text(x)
-
-  // Core Expressions.
-  def read(k: Transaction): Transaction = Expression.read(new Read(k))
-  def write(k: Transaction, v: Transaction): Transaction = Expression.write(new Write(k, v))
-  def load(n: Transaction): Transaction = Expression.load(new Load(n))
-  def store(n: Transaction, v: Transaction): Transaction = Expression.store(new Store(n, v))
-  def branch(c: Transaction, p: Transaction, f: Transaction): Transaction = Expression.branch(new Branch(c, p, f))
-  def cons(a: Transaction, b: Transaction): Transaction = Expression.cons(new Cons(a, b))
-  def repeat(c: Transaction, b: Transaction): Transaction = Expression.repeat(new Repeat(c, b))
-  def prefetch(k: Transaction): Transaction = Expression.prefetch(new Prefetch(k))
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/scala/Record.scala

@@ -1,544 +1,0 @@
-//package caustic.syntax.core
-//
-///**
-// *
-// * @param key
-// */
-//case class Record(key: Transaction)
-//    extends Evaluable
-//    with Assignable
-//    with Deletable
-//    with Chainable[Record] {
-//
-//  /**
-//   *
-//   * @return
-//   */
-//  def kind: Transaction = add(this.key, Record.Field, "__kind__")
-//
-////  /**
-////   *
-////   * @param to
-////   * @return
-////   */
-////  def copy(to: Transaction): Transaction =
-////    cons(
-////      // Construct a stack to recursively process fields of a structure.
-////      store("__stack__", add(this.key, Record.Array)),
-////      store("__prefix__", to),
-////
-////      repeat(ne(load("__stack__"), Empty), cons(
-////        // Pop the head of the stack and its kind.
-////        store("__i__", indexOf(load("__stack__"), Record.Array)),
-////        store("__head__", slice(load("__stack__"), Zero, load("__i__"))),
-////        store("__kind__", add(load("__head__"), Record.Field, "__kind__")),
-////        store("__stack__", slice(load("__stack__"), add(load("__i__"), length(Record.Array)))),
-////
-////        // Recurse on stuctures.
-////        branch(equal(read(load("__kind__")), "structure"), cons(
-////          // Prefix all fields with the head key.
-////          store("__fields__", read(load("__head__"))),
-////          store("__i__", Zero),
-////
-////          repeat(less(load("__i__"), length(load("__fields__"))), cons(
-////            store("__next__", indexOf(slice(load("__fields__"), load("__i__")), Record.Array)),
-////            store("__prefix__", slice(load("__fields__"), Zero, load("__i__"))),
-////            store("__suffix__", slice(load("__fields__"), load("__next__"))),
-////
-////            store("__name__", slice(load("__fields__"), load("__i__"), load("__next__"))),
-////            store("__key__", add(load("__head__"), Record.Field, load("__name__"))),
-////            store("__fields__", add(load("__prefix__"), load("__key__"), load("__suffix__"))),
-////            store("__i__", add(add(load("__next"), length("__key__")), length(Record.Array)))
-////          )),
-////
-////          // Prefetch keys and append them to the stack.
-////          prefetch(load("__fields__")),
-////          store("__stack__", add(load("__stack__"), load("__fields__"), Record.Array))
-////        )),
-////
-////        // Prepend the new field with the new key.
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/scala/ScalaCompiler.scala

@@ -1,95 +1,0 @@
-package caustic.syntax.scala
-
-import caustic.grammar.CausticParser
-import caustic.grammar.CausticParser.DeclarationContext
-import caustic.syntax.Compiler
-
-import org.antlr.v4.runtime.misc.ParseCancellationException
-
-import scala.collection.JavaConverters._
-import scala.collection.mutable
-
-/**
- *
- */
-case class ScalaCompiler(
-  symbols: mutable.Map[String, Type]
-) extends Compiler {
-
-  override def visitProgram(ctx: CausticParser.ProgramContext): String = {
-    // Extract the module.
-    val module = ctx.module(0).getText
-
-    // Extract the imports, and compile the files.
-    val imports = ctx.module().asScala.drop(1)
-      .map(m => s"""import ${m.getText}""")
-      .mkString("\n")
-
-    // Extract the declarations.
-    val declarations = ctx.declaration().asScala
-      .map(visitDeclaration)
-      .mkString("\n")
-
-    // Serialize the module and imports.
-    s"""package $module
-       |$imports
-       |$declarations
-     """.stripMargin
-  }
-
-  override def visitDeclaration(ctx: DeclarationContext): String =
-    super.visitChildren(ctx)
-
-  override def visitService(ctx: CausticParser.ServiceContext): String =
-    super.visitChildren(ctx)
-
-  override def visitFunction(ctx: CausticParser.FunctionContext): String = {
-    // Extract the function name.
-    val name = ctx.Identifier(0).getText
-
-    // Extract the function arguments.
-    val args = ctx.parameters().parameter().asScala
-      .map(x => (x.Identifier(0).getText, x.Identifier(1).getText))
-      .map(x => Parameter(x._1, this.symbols(x._2)))
-
-    // Extract the function return value.
-    val returns = ctx.Identifier(1).getText
-
-    // Insert the record declaration into the symbol table.
-    this.symbols += ctx.Identifier(0).getText -> Function(this.symbols(returns), args.toList)
-
-    s"""def $name(${visitParameters(ctx.parameters())}): $returns = {
-       |
-       |}
-     """.stripMargin
-  }
-
-  override def visitRecord(ctx: CausticParser.RecordContext): String = {
-    // Extract the record name and fields.
-    val name = ctx.Identifier(0).getText
-    val fields = ctx.parameters().parameter().asScala
-      .map(x => (x.Identifier(0).getText, x.Identifier(1).getText))
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/scala/symbols.scala

@@ -1,17 +1,0 @@
-package caustic.syntax.scala
-
-/**
- *
- */
-sealed trait Type
-case class Primitive(name: String) extends Type
-case class Function(args: Map[String, Type], returns: String, result: String) extends Type
-case class Record(fields: Map[String, Type]) extends Type
-case class Service(functions: Map[String, Function]) extends Type
-
-/**
- *
- */
-sealed trait Symbol
-case class Variable(name: String, of: Type)
-case class Reference(key: String, to: Type)

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/scala/visitors/BlockVisitor.scala

@@ -1,85 +1,0 @@
-package caustic.syntax.scala
-package visitors
-
-import caustic.grammar.CausticBaseVisitor
-import caustic.grammar.CausticParser._
-
-/**
- *
- * @param namespace
- * @param symbols
- */
-case class BlockVisitor(
-  symbols: Map[String, Symbol]
-) extends CausticBaseVisitor[Transaction] {
-
-  override def visitBlock(ctx: BlockContext): Transaction =
-    super.visitBlock(ctx)
-
-  override def visitStatement(ctx: StatementContext): Transaction =
-    super.visitStatement(ctx)
-
-  override def visitRollback(ctx: RollbackContext): Transaction =
-    super.visitRollback(ctx)
-
-  override def visitDefinition(ctx: DefinitionContext): Transaction =
-    super.visitDefinition(ctx)
-
-  override def visitAssignment(ctx: AssignmentContext): String = {
-    // Update the left-hand side with the right-hand side.
-    this.symbols(ctx.symbol()) match {
-      case Variable(name, _) => s"store(add(load(text("__scope__"), text($name)), $rhs)"
-      case Reference(key, _) => s"write(text($key), $rhs)"
-    }
-
-
-    // Determine the right-hand side of the assignment operation.
-    val rhs = ctx match {
-      case _ if ctx.Assign() != null =>
-        visitResult(ctx.result())
-      case _ if ctx.AddAssign() != null =>
-        add(this.expression.visitSymbol(ctx.symbol()), visitResult(ctx.result()))
-      case _ if ctx.SubAssign() != null =>
-        sub(this.expression.visitSymbol(ctx.symbol()), rhs)
-      case _ if ctx.MulAssign() != null =>
-        mul(this.expression.visitSymbol(ctx.symbol()), rhs)
-      case _ if ctx.DivAssign() != null =>
-        div(this.expression.visitSymbol(ctx.symbol()), rhs)
-      case _ if ctx.ModAssign() != null =>
-        mod(this.expression.visitSymbol(ctx.symbol()), rhs)
-    }
-
-    // Determine the left-hand and right-hand side of the assignment.
-    val lhs = visitSymbol(ctx.symbol())
-
-
-    super.visitChildren(ctx)
-  }
-
-  override def visitValue(ctx: ValueContext): Transaction = {
-    if (ctx.expression() != null)
-      ExpressionVisitor(this.symbols).visitExpression(ctx.expression())
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/scala/visitors/ServiceVisitor.scala

@@ -1,29 +1,0 @@
-package caustic.syntax.scala.visitors
-
-import caustic.grammar.CausticBaseVisitor
-import caustic.grammar.CausticParser.{DeclarationContext, FunctionContext, ParameterContext, ParametersContext, RecordContext, ServiceContext}
-
-/**
- *
- */
-class ServiceVisitor extends CausticBaseVisitor[String] {
-
-  override def visitService(ctx: ServiceContext): String =
-    super.visitService(ctx)
-
-  override def visitDeclaration(ctx: DeclarationContext): String =
-    super.visitDeclaration(ctx)
-
-  override def visitRecord(ctx: RecordContext): String =
-    super.visitRecord(ctx)
-
-  override def visitFunction(ctx: FunctionContext): String =
-    super.visitFunction(ctx)
-
-  override def visitParameters(ctx: ParametersContext): String =
-    super.visitParameters(ctx)
-
-  override def visitParameter(ctx: ParameterContext): String =
-    super.visitParameter(ctx)
-
-}

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/tcompiler/ThriftCompiler.scala

@@ -1,92 +1,0 @@
-package caustic.syntax
-package tcompiler
-
-import caustic.grammar.CausticParser._
-import caustic.syntax.tcompiler.ThriftCompiler._
-
-import org.antlr.v4.runtime.misc.ParseCancellationException
-import scala.collection.JavaConverters._
-
-/**
- * A compiler that generates a Thrift RPC interface. For some foo.acid, the ThriftCompiler generates
- * a Thrift IDL foo.thrift, and an executable FooThriftServer.scala which serves the Caustic
- * implementation when run. The ThriftCompiler gives Caustic out-of-box interoperability with a
- * variety of different programming languages and software stacks.
- *
- * @param namespace Thrift namespace.
- */
-case class ThriftCompiler(namespace: String) extends Compiler {
-
-  override def visitService(ctx: ServiceContext): String = {
-    // Recurse on all declarations.
-    val declarations = ctx.declaration().asScala.map(visitDeclaration)
-    val structs = declarations.filter(_ startsWith "record").mkString("\n")
-    val definitions = declarations.filter(_ startsWith "def").mkString("\n")
-
-    // Construct the Thrift IDL.
-    s"""namespace * ${this.namespace}
-       |$structs
-       |
-       |service ${ctx.Identifier().getText} {
-       |$definitions
-       |
-       |}
-     """.stripMargin
-  }
-
-  override def visitDeclaration(ctx: DeclarationContext): String = {
-    if (ctx.record() != null) {
-      val name = ctx.record().Identifier().getText
-      val params = serialize(ctx.record().parameters())
-
-      // Serialize records as Thrift structs.
-      s"""
-         |struct $name {
-         |$params
-         |}
-       """.stripMargin
-    } else if (ctx.function() != null) {
-      // Determine the Thrift return type of the function.
-      val isref = ctx.function().Ampersand() != null
-      val rtype = ctx.function().Identifier(1).getText
-      val ttype = (isref, rtype) match {
-        case (true, _) => "string"
-        case (_, "Boolean") => "bool"
-        case (_, "Integer") => "i64"
-        case (_, "Decimal") => "double"
-        case (_, "String") => "string"
-        case (_, x) => x
-      }
-
-      // Extract the name of the function and indent its Thrift parameters.
-      val name = ctx.function().Identifier(0)
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/visitors/BlockVisitor.scala

@@ -1,0 +1,118 @@
+package caustic.syntax
+package visitors
+
+import caustic.grammar.CausticBaseVisitor
+import caustic.grammar.CausticParser._
+
+import scala.collection.JavaConverters._
+import scala.collection.mutable
+
+/**
+ *
+ * @param namespace
+ * @param symbols
+ */
+case class BlockVisitor(
+  symbols: Map[String, Symbol],
+  namespace: String
+) extends CausticBaseVisitor[String] {
+
+  // Local symbol table.
+  val locals: mutable.Map[String, Symbol] = mutable.Map.empty
+
+  // Monotonically increasing label number.
+  var label: Int = 0
+
+  /**
+   * Returns the result of visiting the specified BlockContext. Visitor is scoped using the
+   * monotonically increasing label number and nested blocks inherit the local symbols of their
+   * parent block.
+   *
+   * @param ctx BlockContext to visit.
+   * @return Result of visiting BlockContext.
+   */
+  def scope(ctx: BlockContext): String = {
+    val frame = this.namespace + "$" + this.label
+    this.label += 1
+    BlockVisitor(this.symbols ++ this.locals, frame).visitBlock(ctx)
+  }
+
+  override def visitBlock(ctx: BlockContext): String =
+    ctx.statement().asScala
+      .map(visitStatement)
+      .foldLeft("Empty")((a, b) => s"""cons($a, $b)""")
+
+  override def visitRollback(ctx: RollbackContext): String =
+    s"""rollback(${ExpressionVisitor(this.symbols).visitExpression(ctx.expression())})"""
+
+  override def visitDefinition(ctx: DefinitionContext): String = {
+    super.visitDefinition(ctx)
+  }
+
+  override def visitAssignment(ctx: AssignmentContext): String = {
+    this.symbols(ctx.variable().Identifier(0).getText) match {
+      case Variable(name, true, Primitive) =>
+        s"""store(text("$name"), ${visitValue(ctx.value)})"""
+      case Variable(name, true, Record(f)) =>
+        // Set every field of the record equal to the provided record.
+
+      case Variable(name, true, Reference(Primitive)) =>
+        s"""write(text("$name", ${visitValue(ctx.value)})"""
+      case Variable(name, true, Reference(Record(f))) =>
+        // Set every field of the referenced record equal to the provided record.
+
+    }
+
+    super.visitChildren(ctx)
+  }
+
+  override def visitValue(ctx: ValueContext): String = {
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/visitors/DeclarationVisitor.scala

@@ -1,0 +1,88 @@
+package caustic.syntax
+package visitors
+
+import caustic.grammar.CausticBaseVisitor
+import caustic.grammar.CausticParser._
+import caustic.runtime.{Flag, Real, Text, Transaction, thrift}
+import org.apache.thrift.TException
+import org.apache.thrift.async.AsyncMethodCallback
+import scala.collection.mutable
+import scala.collection.JavaConverters._
+import scala.util.{Failure, Success}
+
+/**
+ *
+ */
+class DeclarationVisitor(
+  symbols: Map[String, Symbol],
+  namespace: String
+) extends CausticBaseVisitor[String] {
+
+  // Local symbol table.
+  val locals: mutable.Map[String, Symbol] = mutable.Map.empty
+
+  override def visitDeclaration(ctx: DeclarationContext): String =
+    super.visitDeclaration(ctx)
+
+  override def visitRecord(ctx: RecordContext): String = {
+    // Extract the service signature.
+    val name = ctx.Identifier().getText
+    val args = visitParameters(ctx.parameters())
+
+    // Add the record to the symbol table.
+
+    // Do not serialize the record.
+
+  }
+
+  override def visitService(ctx: ServiceContext): String = {
+    // Extract the service signature.
+    val name = ctx.Identifier()
+    val functions = ctx.function().asScala.map(visitFunction).mkString("\n")
+
+    // Add the service to the symbol table.
+
+    // Serialize the service.
+    s"""case class $name(database: thrift.Database.Client) extends thrift.$name {
+       |$functions
+       |}
+     """.stripMargin
+  }
+
+  override def visitFunction(ctx: FunctionContext): String = {
+    // Extract the function signature.
+    val name = ctx.Identifier().getText
+    val args = visitParameters(ctx.parameters())
+    val ret  = visitType(ctx.`type`())
+
+    // Visit the body of the function.
+    val body = BlockVisitor(this.symbols ++ this.locals, this.namespace + "$" + name)
+      .visitBlock(ctx.block())
+
+    // Add the function to the symbol table.
+
+    // Parse the output of the body into the return value.
+    val result =
+
+    // Convert the function into executable Scala.
+    s"""  override def $name($args, res: AsyncMethodCallback[$ret]): Unit = {
+       |    this.database.execute($body).onComplete {
+       |      case Success(_) => res.onComplete($result)
+       |      case Failure(e: TException) => res.onError(e)
[…]

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/visitors/ExpressionVisitor.scala

@@ -1,12 +1,14 @@
-package caustic.syntax.scala
+package caustic.syntax
 package visitors
 
-import caustic.grammar.CausticBaseVisitor
+import caustic.grammar._
 import caustic.grammar.CausticParser._
+import scala.collection.JavaConverters._
 
 /**
+ * An expression parser.
  *
- * @param symbols
+ * @param symbols Symbol table.
  */
 case class ExpressionVisitor(
   symbols: Map[String, Symbol]
@@ -20,25 +22,58 @@ case class ExpressionVisitor(
     else if (ctx.Number() != null)
       s"""real(${ctx.Number().getText.toDouble})"""
     else if (ctx.String() != null)
-      s"""text(${ctx.String().getText})"""
+      s"""text("${ctx.String().getText}")"""
     else
       visitChildren(ctx)
   }
 
-  override def visitSymbol(ctx: SymbolContext): String = this.symbols(ctx.getText) match {
-    case x: Variable => s"""load(add(load(text("__scope__"), text(${x.name})))"""
-    case x: Reference => s"""read(text(${x.key}))"""
-    case _ => super.visitChildren(ctx)
+  override def visitVariable(ctx: VariableContext): String = {
+    // Determine field of the specified record and automatically dereference pointers.
+    def field(parent: Type, names: List[String], key: String): String = (parent, names) match {
+      case (Primitive, Nil) =>
+        key
+      case (Reference(Primitive), Nil) =>
+        s"""read($key)"""
+      case (Record(f), head :: rest) =>
+        field(f(head), rest, s"""add($key, text("."), text("$head"))""")
+      case (Reference(Record(f)), head :: rest) =>
+        field(f(head), rest, s"""add(read($key), text("."), text("$head"))""")
+    }
+
+    // Load the corresponding variable.
+    this.symbols(ctx.Identifier(0)) match {
+      case Variable(name, _, Primitive) =>
+        // Load the value of primitive variables.
+        s"""load(text("$name"))"""
+      case Variable(name, _, record: Record) =>
+        // Load the corresponding field of the record.
+        val names = ctx.Identifier().asScala.drop(1).map(_.getText).toList
+        val key = field(record, names, s"""text("$name")""")
+        s"""load($key)"""
+      case Variable(name, _, Reference(Primitive)) =>
[…]

### 2017-09-24 caustic-syntax/src/test/scala/caustic/syntax/LanguageTest.scala

@@ -1,16 +1,0 @@
-package caustic.runtime
-package syntax
-
-import org.junit.runner.RunWith
-import org.scalatest.concurrent.ScalaFutures
-import org.scalatest.mockito.MockitoSugar
-import org.scalatest.{AsyncFunSuite, Matchers}
-import org.scalatest.junit.JUnitRunner
-
-@RunWith(classOf[JUnitRunner])
-class LanguageTest extends AsyncFunSuite
-  with ScalaFutures
-  with MockitoSugar
-  with Matchers {
-
-}

### 2017-09-24 caustic-syntax/src/test/scala/caustic/syntax/RecordTest.scala

@@ -1,15 +1,0 @@
-package caustic.syntax
-
-import org.junit.runner.RunWith
-import org.scalatest.{FunSuite, Matchers}
-import org.scalatest.concurrent.ScalaFutures
-import org.scalatest.junit.JUnitRunner
-import org.scalatest.mockito.MockitoSugar
-
-@RunWith(classOf[JUnitRunner])
-class RecordTest extends FunSuite
-  with ScalaFutures
-  with MockitoSugar
-  with Matchers {
-
-}

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/Cache.scala
Fixed compile errors

@@ -78,6 +78,9 @@ trait Cache extends Database {
       case Failure(e: ConflictException) =>
         // Invalidate cached keys to force them to reload from the database.
         invalidate(e.keys.asScala.toSet).transform(_ => Failure(e))
+      case Failure(_) =>
+        // Ignore all other failures.
+        Future.unit
     }
 
 }

### 2017-09-24 caustic-runtime/src/main/scala/caustic/runtime/relational/PostgresDatabase.scala

@@ -13,7 +13,7 @@ import scala.concurrent.ExecutionContext
  * @param underlying Underlying PostgreSQL DataSource.
  * @param ec Implicit execution context.
  */
-case class PostgreSQLDatabase(underlying: DataSource)(
+case class PostgresDatabase(underlying: DataSource)(
   implicit ec: ExecutionContext
 ) extends RelationalDatabase(underlying) {
 

### 2017-09-24 caustic-runtime/src/test/scala/caustic/runtime/memory/LocalDatabaseTest.scala

@@ -1,7 +1,6 @@
 package caustic.runtime
 package memory
 
-import caustic.runtime.databases.LocalDatabase
 import org.junit.runner.RunWith
 import org.scalatest.Outcome
 import org.scalatest.junit.JUnitRunner

### 2017-09-24 caustic-runtime/src/test/scala/caustic/runtime/relational/MySQLDatabaseTest.scala

@@ -1,11 +1,11 @@
 package caustic.runtime
 package relational
 
-import caustic.runtime.relational.MySQLDatabase
 import com.mchange.v2.c3p0.ComboPooledDataSource
 import org.junit.runner.RunWith
 import org.scalatest.junit.JUnitRunner
 import org.scalatest.{BeforeAndAfterAll, Outcome}
+import scala.concurrent.ExecutionContext.Implicits.global
 
 @RunWith(classOf[JUnitRunner])
 class MySQLDatabaseTest extends DatabaseTest with BeforeAndAfterAll {

### 2017-09-24 caustic-runtime/src/test/scala/caustic/runtime/relational/PostgresDatabaseTest.scala

@@ -1,10 +1,11 @@
-package caustic.runtime.relational
+package caustic.runtime
+package relational
 
-import caustic.runtime.DatabaseTest
 import com.mchange.v2.c3p0.ComboPooledDataSource
 import org.junit.runner.RunWith
 import org.scalatest.junit.JUnitRunner
 import org.scalatest.{BeforeAndAfterAll, Outcome}
+import scala.concurrent.ExecutionContext.Implicits.global
 
 @RunWith(classOf[JUnitRunner])
 class PostgresDatabaseTest extends DatabaseTest with BeforeAndAfterAll {
@@ -27,7 +28,7 @@ class PostgresDatabaseTest extends DatabaseTest with BeforeAndAfterAll {
     con.close()
 
     // Run the tests.
-    val database = PostgreSQLDatabase(this.pool)
+    val database = PostgresDatabase(this.pool)
     test(database)
   }
 

### 2017-09-24 caustic-syntax/src/main/scala/caustic/syntax/ThriftCompiler.scala

@@ -69,7 +69,7 @@ class ThriftCompiler extends Compiler {
       case "Integer" => "i64"
       case "Decimal" => "double"
       case "String"  => "string"
-      case _ => _
+      case x => x
     }
 
 }

### 2018-04-04 caustic-library/src/main/scala/caustic/library/control/Context.scala
Renamed Null to None

@@ -1,7 +1,7 @@
 package caustic.library.control
 
 import caustic.library.typing._
-import caustic.runtime
+import caustic.runtime._
 
 /**
  * A parsing context.
@@ -9,7 +9,7 @@ import caustic.runtime
  * @param body Current program.
  */
 case class Context(
-  private[library] var body: runtime.Program = Null,
+  private[library] var body: Program = None,
   private[library] var current: scala.Int = -1
 ) {
 
@@ -18,8 +18,8 @@ case class Context(
    *
    * @param that Program to append.
    */
-  def +=(that: runtime.Program): Unit =
-    this.body = runtime.cons(this.body, that)
+  def +=(that: Program): Unit =
+    this.body = cons(this.body, that)
 
   /**
    *

### 2018-04-04 caustic-library/src/main/scala/caustic/library/record/ops/delete.scala

@@ -28,7 +28,7 @@ object delete extends Poly2 {
     selector: Selector.Aux[TRepr, FieldName, FieldType],
     field: Field.Aux[FieldType, Variable[FieldType]]
   ): Case.Aux[Args[T], FieldName, Args[T]] = at[Args[T], FieldName] { (x, f) =>
-    field(x.src, f.name).set(Null)
+    field(x.src, f.name).set(None)
     x
   }
 
@@ -51,7 +51,7 @@ object delete extends Poly2 {
     evidence: FieldType <:< Reference[FieldT]
   ): Case.Aux[Args[T], FieldName, Args[T]] = at[Args[T], FieldName] { (x, f) =>
     if (x.recursive) field(x.src, f.name).delete(x.recursive)
-    x.src.pointer.set(Null)
+    x.src.pointer.set(None)
     x
   }
 

### 2018-04-04 caustic-library/src/main/scala/caustic/library/record/package.scala

@@ -1,6 +1,7 @@
 package caustic.library
 
 import caustic.library.control.Context
+
 import shapeless.{HList, LabelledGeneric}
 import shapeless.ops.hlist.LeftFolder
 import shapeless.ops.record.Keys

### 2018-04-04 caustic-library/src/main/scala/caustic/library/typing/package.scala

@@ -8,7 +8,7 @@ import scala.language.implicitConversions
 
 package object typing {
 
-  val Null                                          : Program           = runtime.Null
+  val None                                          : Program           = runtime.Null
   val True                                          : Value[Boolean]    = Constant(true)
   val False                                         : Value[Boolean]    = Constant(false)
 
@@ -72,11 +72,11 @@ package object typing {
   }
 
   implicit class ToJsonOps[T <: Primitive](x: Value[T]) {
-    def toJson: Value[String] = branch(x <> Null, x, "null")
+    def toJson: Value[String] = branch(x <> None, x, "null")
   }
 
   implicit class ToJsonStringOps(x: Value[String]) {
-    def toJson: Value[String] = branch(x <> Null, x.quoted, "null")
+    def toJson: Value[String] = branch(x <> None, x.quoted, "null")
   }
 
 

### 2018-04-06 caustic-assets/paper/main.tex
Updated paper

--- 2017-09-16 caustic-assets/paper/main.tex
@@
 \usepackage{amsthm}
+\usepackage[backend=biber]{biblatex}
+\usepackage{listings}
 \usepackage{hyperref}
+\usepackage{xcolor}
 
-\begin{document}
+% Configuration.
+\lstset{
+  backgroundcolor=\color{white!99!black},
+  numbers=left, numberstyle=\tiny, stepnumber=1, numbersep=5pt,
+}
+
+\lstdefinestyle{Caustic}{
+  basicstyle=\small\ttfamily\color{black},
+  breaklines=true,
+  moredelim=[s][\color{green!50!black}\ttfamily]{'}{'},
+  moredelim=*[s][\color{black}\ttfamily]{options}{\}},
+  commentstyle={\color{gray}\itshape},
+  morecomment=[l]{//},
+  emph={def, del, elif, else, extends, if, import, module, record, return, rollback, service, while, var},
+  emphstyle={\color{blue}\ttfamily},
+  alsoletter={:,|,;},
+  morekeywords={:,|,;},
+  keywordstyle={\color{black}},
+}
+
+\hypersetup{
+  colorlinks=true,
+  linkcolor=black,
+  citecolor=black,
+  filecolor=blue,
+  urlcolor=blue
+}
+
+% Bibliography.
+\bibliography{main}
+\nocite{*}
 
 % Title.
-\title{Caustic: Reinventing Database Transactions}
+\begin{document}
+\title{Caustic: A Transactional Programming Language}
 \author{Ashwin Madavan}
@@
 \begin{abstract}
-Database languages are designed for ease of implementation and not ease of use.
-For example, every relational database has its own interpretation of the SQL specification that is most convenient for the underlying database architecture and most often incompatible with its competition, and most key-value stores expose primtiive interfaces that are lacking in functionality and programmability.
-Caustic is a robust query language that is intuitive to use and straightforward to integrate with any database.
+Many programming languages provide fundamental abstractions such as locks, semaphores, and monitors
+to explicitly deal with race conditions. Some, like Rust \cite{rust}, go a step further and are able
+to statically detect race conditions between concurrent threads. But none, however, are able
+guarantee that race conditions will be completely eliminated from distributed systems. Distributed
+systems form the computing backbone of nearly every major technology from social networks to video
+streaming, but their intricate complexity coupled with the inability to detect race conditions makes
+designing them extremely error-prone. Caustic is a programming language for building correct
+distributed systems. Programs written in Caustic may be distributed arbitrarily without the use of
+any explicit synchronization, and will never exhibit race conditions.
 \end{abstract}
@@
 \section{Introduction}
-The syntax of query languages varies wildly between databases.
-For example, relational databases like MySQL and PostgreSQL claim to implement the same SQL specification, but, in reality, all but the most trivial of queries are incompatible with each other.
-This lack of uniformity between languages tightly couples the database with the queries that are run on it, and makes it difficult to switch storage systems as requirements change.
+Concurrency is \emph{hard}. Concurrency refers to situations in which multiple programs
+simultaneously modify shared data. Concurrent programs may be run across threads, processes,
+and, in the case of distributed systems, networks. Concurrency is challenging because it introduces
+ambiguity in execution order, and it is precisely this ambiguity that causes race conditions. Race
+conditions are situations in which the execution order of concurrent programs affects the outcome.
+For example, suppose there exist two programs $A$ and $B$ that each increment a shared counter $x$.
+Each program must first read the current value of $x$ in order to write $x + 1$. If $B$ reads
+\emph{after} $A$ writes, then $B$ reads $x + 1$ and writes $x + 2$. However, if $B$ reads
+\emph{before} $A$ writes but after $A$ reads, then both $A$ and $B$ will read $x$ and write $x + 1$.
+This is an example of a race condition, because the value of the counter $x$ depends on the order in
+which $A$ and $B$ are executed. This race condition may seem relatively benign, but it can have
+catastrophic consequences in practical systems. Suppose the value of $x$ corresponded to your bank
+balance. What if your bank determined your balance differently depending on the order in which
+deposits are made? Race conditions manifest themselves in subtle ways in concurrent systems, and
+they can often be difficult to detect and challenging to remove.
 
-The structure of query languages is deeply motivated by the capabilities and limits of the underlying database and not by what applications actually require.
-In particular, the language of relational algebra, which SQL is based upon, is not first-nature for most programmers.
-[On the Principles of Good Language Design, CJ Date] [http://wiki.c2.com/?SqlFlaws]
-
-It's clear that query languages are typically designed for ease of implementation and not ease of use.
-They have disparate interfaces, deficient functionality, and unintuitive syntax.
-Caustic provides both a robust language for specifying transactions and an efficient runtime for executing them on arbitrary databases.
-
-% Background
-\section{Background}
-
-% Runtime
-\section{Runtime}
-The runtime is responsible for executing transactions on arbitrary databases.
-Transactions are represented as abstract-syntax trees, because they are easy to manipulate, serialize, and optimize.
-The runtime exposes a Thrift interface for cross-language transaction execution, and is capable of running on any database that supports the following primitive operations:
+\subsection{Eliminating Race Conditions}
+Before introducing mechanisms to deal with race conditions, we must first define the necessary and
+sufficient criteria that any such mechanism must satisfy to correctly mitigate race conditions.
+Race conditions are a relatively well-studied problem in database literature, and we may draw upon
+known results from the field to define four properties of race-free programs. \cite{transactions}
 
 \begin{itemize}
-  \item \textbf{get}: Returns the values of a set of keys.
-  \item \textbf{cput}: Conditionally apply changes iff dependent keys remain unchanged.
+\item \textbf{Atomic}: Programs are all-or-nothing; they are never partially applied.
+\item \textbf{Consistent}: Programs see the effect of all completed programs.
+\item \textbf{Isolated}: Programs cannot see the effect of in-progress programs.
+\item \textbf{Durable}: The effects of completed programs are permanently visible.
 \end{itemize}
 
-\subsection{Representation}
+\subsection{Locks and Leases}
+A common approach to dealing with race conditions is locking. Before a program accesses or
+modifies shared data, it first declares its intentions to other programs by acquiring a
+lock that it subsequently releases when it is finished with the data. Locks trivially satisfy
+the ACID criteria, because they require programs to have exclusive ownership before performing any
+operations on shared data. However, locking introduce new problems.
 
-\subsection{Snapshot Isolation}
+First, concurrent acquisition of multiple locks can cause a deadlock that prevents the system from
+making progress. For example, if program $A$ acquires lock $x$ and then attempts to acquire $y$
+and program $B$ acquires $y$ and then attempts to acquire $x$, then neither $A$ nor $B$ can make
+progress because each is waiting for the other to release their lock. This problem is typically
+mitigated by imposing a total order on lock acquisition; for any two locks $x$ and $y$, $x$ will
+always be acquired before $y$ or $y$ will always be acquired before $x$.
 
-\subsection{Optimizations}
-Because transactions are represented as abstract-syntax trees, the runtime is able to perform some interesting optimizations to improve execution performance.
-Most importantly, the runtime performs read and write batching.
-In other words, the runtime fetches \emph{as many reads as possible} and applies \emph{all writes together}.
-Clearly, the runtime performs optimally few roundtrips to and from the underlying key-value store, which is typically the performance bottleneck in database applications.
+Second, faulty programs may never release their locks. For example, if program $A$ acquires lock
+$x$ and subsequently fails, then no other program can ever acquire $x$. This problem is typically
+mitigated by using leases. \cite{leases} A lease is a lock that is automatically released after a
+certain amount of time. Programs must be carefully constructed so that they complete their
+operations on shared data within the lease duration or refresh the lease before it expires.
 
-\subsection{Tunable Consistency}
-Transactions are executed using MVCC, and derive their consistency and isolation guarantees from the underlying key-value store.
-If the underlying key-value has a repeatable isolation level, for example, then so will transactions in Caustic.
-If the underlying key-value is not transactional, then neither are transactions in Caustic.
-Therefore, applications may substitute a key-value store with weaker transactional guarantees (or no transactions at all!) to improve performance.
+Third, locking is expensive. Locks must be acquired regardless of whether or not there actually
+are concurrent operations on shared data, because programs cannot know if there are or will be
+other programs that want to simultaneously use the data. This significantly degrades
+performance in situations where contention between programs is low.
 
-\subsection{Benchmarks}
-Compare the Caustic runtime (running on RocksDB) to comparable systems like Calvin and Tango.
-Rely on quantitative metrics like end-to-end latency and transaction throughput on standard database benchmarks like YCSB and TPC-C.
+Fourth, locking cannot protect against programmer error. Programmers may omit or incorrectly use a
+lock and thereby introduce race conditions into their program. Locks cannot guarantee they will be
+used correctly, and, therefore, cannot guarantee that race conditions will be exhaustively
+eliminated from a program.
 
-\subsubsection{Effect of Transaction Size}
-Performance scales linearly with transaction size. This is enabled by the tail recursive design of the runtime and can be verified empirically from benchmarks.
+\subsection{Database Transactions}
+An alternative approach is to use a transactional database to protect shared data from concurrent
+modification. There are a number of storage systems that provide ACID transactions in full or in
+limited capacity. However, each of these storage systems has their own bespoke interface for
+specifying transactions that are often lacking in functionality and performance. Recent years have
+marked a proliferation in NoSQL databases that scale well by shedding functionality. These
+databases were not popularized because of their query languages, they were \emph{in spite} of
+them. Some, like Cassandra and Aerospike, attempt to mimic the relational semantics of SQL, but
+they fall short of implementing the entire SQL specification. Others, like MongoDB and DynamoDB,
+implement entire new query languages. Even SQL is not beyond reproach. SQL lacks a canonical
+implementation and has ambiguous and unintuitive syntax. \cite{sql} Relational databases like
+MySQL and PostgreSQL that each claim to implement the same SQL specification actually implement
+incompatible subsets of its functionality. Stack differences between query languages tightly
+couples storage systems and the programs that are run on them, and makes the choice of database
+in an application effectively permanent. While transactional storage systems provide the necessary
+guarantees on which correct distributed systems can be built, their collective lack of a robust
+and uniform interface makes it all but impossible to design non-trivial applications.
 
-\subsubsection{Effect of Read or Write Skew}
-Transactions with more reads than writes will have a different performance profiles than transactions with more writes than reads.
-However, this difference in performance is an artifact of the choice of underlying key-value store and not due to a read or write bias in the runtime.
-In other words, a read optimized key-value store will fare better with read-heavy transactions and a write optimized key-value store with write-heavy transactions in Caustic.
+\subsection{Optimistic Concurrency}
+Optimistic concurrency allows multiple programs to simultaneously access, but not modify, shared
+data without acquiring locks. Each program locally buffers any modifications that it makes and
+attempts to atomically apply the modifications when it completes conditioned on the data that it
+accessed remaining unchanged. If any data was modified, the program retries. This conditional
+update, known as a multi-word compare-and-swap and referred to as a \textbf{transaction}, is known
+to satisfy the ACID criteria and is widely used in a number of software transactional memory
+systems including Caustic. \cite{stm} Optimistic concurrency assumes that contention between
+programs will be low, because frequent retries can significantly degrade performance.
 
-\subsubsection{Effect of Contention}
-What happens to performance under contention?
-The number of retries required to execute a transaction may be modeled as a negative binomial distribution in which the probability of success $p$ is the contention probability.
-Therefore, $A = 1 + NB(1, p)$ is the distribution of attempts.
-Known results about the negative binomial distribution, may be used to make predictions about the mean and variance of transaction execution latency under contention.
-For example, the mean number of attempts $\bar{A} = 1 + \frac{p}{1 - p} = \frac{1}{1 - p}$.
-The contention probability is determined by a number of factors including: the number of keys that are read and written, the number of database reads and writes, and the latency of database reads and write.
+\section{Architecture}
+Caustic is composed of three components: a \textbf{runtime} that executes programs, a
+\textbf{standard library} that simplifies program construction, and a \textbf{compiler} that
+simplifies program syntax. In this section, we'll describe each of these components in detail.
 
-% Syntax
-\section{Syntax}
-The Caustic syntax is a new high-level programming language that compiles into transactions on the Caustic runtime.
-Regardless of the features they might have (objects, functions, types, etc.), all programs eventually compile down into load, stores, and operations on memory - and memory is a key-value store.
-The purpose of Caustic syntax is to provide a rich language for expressing \emph{transactional} loads, stores, and operations on \emph{key-value stores}.
+\subsection{Runtime}
+The runtime is a virtual machine that dynamically translates \textbf{programs} into transactions.
+A program is an abstract-syntax tree that is composed of \textbf{literals} and
+\textbf{expressions}. A literal is a scalar value of type flag, real, text, and null which
+correspond to bool, double, string, and null respectively in most C-style languages. An expression
+is a function that transforms literal arguments into a literal result. Expressions may be chained
+together arbitrarily to form complex programs. Table \ref{table:expressions} enumerates the
+various expressions supported by the runtime.
 
-\subsection{Concurrency}
-Because all programs written in Caustic compile into transactions, all programs are automatically thread-safe.
-No fancy synchronization mechanisms are required to build parallelizable programs in Caustic, everything is concurrent by default.
+\subsubsection{Execution}
+The runtime uses iterative partial evaluation to gradually reduce programs into a single literal
+result. Modifications to keys are tracked using multiversion concurrency control. Each key is
+associated with a \textbf{revision}, or versioned value, whose version number is incremented
+each time that its value is changed. A revision $A$ \emph{conflicts with} $B$ if $A$ and $B$
+correspond to the same key and the version of $A$ is less than $B$. Conflict is an asymmetric
+relation; if $A$ conflicts with $B$, then $B$ does not conflict with $A$.
 
-\subsection{Separation of Code and Data}
-Every database has its own language for writing stored procedures.
-Even SQL databases typically support very different subsets of the SQL specification, that are not interoperable for anything but the most trivial of programs.
-Because the syntax varies so wildly between different databases, stored procedures are very tightly coupled with the choice of underlying database.
-This makes it tremendously expensive to change databases once development is under way, and impossible to test stored procedures in isolation.
-Caustic provides a uniform interface over arbitrary databases.
-In other words, Caustic programs written for ione database will be identical to the same program written for any another.
-This allows programmers to select whichever database is most appropriate for their use-case (eg.
-in-memory for tests, MySQL for production).
+The runtime \emph{executes} programs on \textbf{volumes}. A volume is a database that supports
+\emph{get} and \emph{cas}. Get retrieves the revisions of a set of keys and cas transactionally
+updates a set of keys if and only if a set of dependent revisions do not conflict with their
+current revisions in the database. Given correct implementations of get and cas, the runtime
+executes programs according to the following procedure.
 
-\subsection{Interoperability}
-Because of the simplicity of the Caustic grammar and runtime, the compiler is able to generate executable binaries that serve Caustic programs over a variety of mediums over any underlying database.
-Most OLTP applications can be divided into three parts: a database connector (eg.
-SQLAlchemy), a set of stored procedures (eg.
-Flask), and an external interface (eg.
-REST, Thrift).
-By specifying their stored procedures in Caustic, programmers can now rely on the Caustic compiler to generate the first and third parts for them.
+\begin{enumerate}
+\item \textbf{Fetch}: Get all keys that are read or written by the program that have not been
+fetched before and add the returned revisions to a local snapshot.
+\item \textbf{Evaluate}: Recursively replace all expressions with literal arguments with their
+corresponding literal result. For example,
 
-\subsection{Benchmarks}
-Compare the Caustic syntax qualitatively to SQL, Tango, and Java.
-Measure differences in code complexity (measured in Cyclomatic or Halstead Complexity), size of executables, and lines of code for a variety of sample use cases (eg.
-implementing ZooKeeper or BookKeeper).
+\[
+\begin{gathered}
+add(real(1), sub(real(0), real(2))) \rightarrow real(-1)
+\end{gathered}
+\]
 
+The result of all writes is saved to a local buffer and the result of all read
+expressions is the latest value of the key in the local buffer or snapshot.
+\item \textbf{Repeat}: Loop until the program is reduced to a single literal. Because all
+expressions with literal arguments return a literal result, all programs will eventually
+reduce to a single literal.
+\item \textbf{Commit}: Cas all keys in the local buffer conditioned on all revisions in
+the local snapshot. Because programs are executed with snapshot isolation, they are
+guaranteed to be serializable. Serializability implies that concurrent execution has the
+same effect as some sequential execution, and, therefore, that program execution will be
+robust against race conditions.
+\end{enumerate}
+
+\subsubsection{Optimizations}
+First, execution is tail-recursive. Therefore, programs may be composed of arbitrarily many
+nested expressions without overflowing the stack frame. This also allows the Scala compiler is
+able to aggressively optimize execution into a tight loop. Second, the runtime batches I/O.
+Reads are performed simultaneously whenever possible and writes are buffered and simultaneously
+committed. By batching I/O, the runtime performs a minimal number of operations on the database.
+This has significant performance implications, because I/O overhead is overwhelmingly the
+bottleneck by many orders of magnitude. \cite{io}
+
+\subsection{Standard Library}
+The runtime provides native support for an extremely limited subset of the operations that
+programmers typically rely on to write programs. The standard library supplements the
+functionality of the runtime by exposing a rich Scala DSL complete with static types, records,
+math, collections, and control flow.
+
+\subsubsection{Typing}
+The runtime natively supports just four dynamic types: $flag$, $real$, $text$, and $null$.
+Dynamic versus static typing is a religious debate among programmers. Advocates of dynamic
+typing often mistakenly believe that type inference and coercive subtyping cannot be provided by
+a static type system. In fact, they can. Because static type systems are able to
+detect type inaccuracies at compile-time, they allow programmers to write more concise and
+correct code. \cite{typing} The standard library provides rich static types and features
+aggressive type inference and subtype polymorphism.
+
+The standard library supports four $Primitive$ types. In descending order of precedence, they
+are: $String$, $Double$, $Int$, $Boolean$. A $Value[+T <: Primitive]$ represents a scalar value.
+Values are covariant in $T$; a $Value[Int]$ is a $Value[Double]$, but a $Value[String]$ is
+not a $Value[Boolean]$. Values may either be $Constant$ or $Variable$. A $Constant$ corresponds
+to an immutable quantity, and a $Variable$ corresponds to a value that is either stored locally
+in memory or remotely in the database.
+
+\subsubsection{Records}
+In addition to these $Primitive$ types, the standard library also supports references to
+user-defined types. A $Reference[T]$ is a reference to an object of type $T$. References use
+\href{https://github.com/milessabin/shapeless}{Shapeless} to materialize compiler macros that
+permit the fields of $T$ to be statically manipulated and iterated. A current limitation is that
+objects cannot be self-referential; an object cannot have a field of its own type.
+
+\subsubsection{Math}
+The runtime natively supports just nine mathematical operations: $add$, $sub$, $mul$, $div$,
+$pow$, $log$, $floor$, $sin$, and $cos$. However, these primitive operations are sufficient to
+derive the entire Scala \href{https://www.scala-lang.org/api/2.12.1/scala/math/index.html}{math}
+library using various mathematical identities and Taylor series approximations. The $div$,
+$log$, $sin$, and $cos$ functions can actually be implemented in terms of the other primitive
+operations; however, native support for them was included in the runtime to improve performance.
+The standard library provides implementations for the functions enumerated in Table
+\ref{table:math}.
+
+\subsubsection{Collections}
+The runtime has no native support for collections of key-value pairs. The standard library
+provides implementations of three fundamental data structures: $List$, $Set$, and $Map$. These
+collections are mutable, statically-typed, and thread-safe. Collections take care of the messy
+details of mapping structured data onto a flat namespace, and feature prefetched iteration. A
+current limitation is that collections may only contain $Primitive$ types.
+
+\subsubsection{Control Flow}
+The runtime has native support for control flow operations like $branch$, $cons$, and $repeat$.
+However, it is syntactically challenging to express these contructs. The standard library
+uses structural types to provide support for $If$, $While$, $Return$, $Assert$, and $Rollback$.
+The standard library uses an implicitly available parsing $Context$ to track modifications made
+to variables, references, and collections and to detect when any control flow statements are
+called.
+
+\subsection{Compiler}
+The standard library provides additional functionality that is deficient in the runtime to make it
+easier to construct programs. However, the standard library does not address the syntactic
+challenges of expressing programs. The compiler translates programs written in the
+statically-typed, object-oriented Caustic programming language into runtime-compatible programs.
+For example, consider the following example of a distributed counter written in Caustic. This
+program may be compiled into a Scala library that may be run without modification on any
+underlying transactional storage volume and distributed arbitrary without error. The Akka project
+also provides an \href{https://git.io/vxS6u}{implementation} of a backend-agnostic distributed
+counter. Their implementation is almost seven times longer.
+
+\begin{lstlisting}[style=Caustic]
+module caustic.example
+
+/**
+ * A count.
+ *
+ * @param value Current value.
+ */
+record Total {
+  value: Int
+}
+
+/**
+ * A distributed counter.
+ */
+service Counter {
+
+ /**
+  * Increments the total and returns its current value.
+  *
+  * @param x Reference to total.
+  * @return Current value.
+  */
+  def increment(x: Total&): Int = {
+    if (x.value) x.value += 1 else x.value = 1
+    x.value
+  }
+
+}
+\end{lstlisting}
+
+\subsection{Generation}
+The compiler uses ANTLR \cite{antlr} to generate a predicated LL(*) parser from an ANTLR grammar
+file. The compiler generates code by walking the parse tree and replacing statements and
+declarations with their equivalent formulation using the standard library. The compiler mangles
+definitions so that they are lexically scoped. It also performs type inference and checking to
+determine and verify static types. The compiler is integrated into the
+\hyperref{https://www.pantsbuild.org/}{Pants} build system and provides a
+\hyperref{https://macromates.com/}{TextMate} bundle that implements syntax highlighting and code
+completion for most text editors and IDEs.
+
+\appendix
+\begin{table}[ht]
+\caption{Runtime Expressions}
+\centering
+\begin{tabular}{l | l}
+\hline\hline
+Expression & Description                                                                          \\
+\hline
+add(x, y)                & Sum of $x$ and $y$.                                                    \\
+both(x, y)               & Bitwise AND of $x$ and $y$.                                            \\
+branch(c, p, f)          & Executes $p$ if $c$ is true, or $f$ otherwise.                         \\
+cons(a, b)               & Executes $a$ and then $b$.                                             \\
+contains(x, y)           & Returns whether or not $x$ contains $y$.                               \\
+cos(x)                   & Cosine of $x$.                                                         \\
+div(x, y)                & Quotient of $x$ and $y$.                                               \\
+either(x, y)             & Bitwise OR of $x$ and $y$.                                             \\
+equal(x, y)              & Returns whether $x$ and $y$ are equal.                                 \\
+floor(x)                 & Floor of $x$.                                                          \\
+indexOf(x, y)            & Returns the index of the first occurrence of $y$ in $x$.               \\
+length(x)                & Returns the number of characters in $x$.                               \\
+less(x, y)               & Returns whether $x$ is strictly less than $y$.                         \\
+load(n)                  & Loads the value of the variable $n$.                                   \\
+log(x)                   & Natural log of $x$.                                                    \\
+matches(x, y)            & Returns whether or not $x$ matches the regex pattern $y$.              \\
+mod(x, y)                & Remainder of $x$ divided by $y$.                                       \\
+mul(x, y)                & Product of $x$ and $y$.                                                \\
+negate(x)                & Bitwise negation of $x$.                                               \\
+pow(x, y)                & Returns $x$ raised to the power $y$.                                   \\
+prefetch(k, s)           & Reads keys at $k/i$ for $i$ in $[0, s)$.                               \\
+read(k)                  & Reads the value of the key $k$.                                        \\
+repeat(c, b)             & Repeatedly executes $b$ while $c$ is true.                             \\
+rollback(r)              & Discards all writes and returns $r$.                                   \\
+sin(x)                   & Sine of $x$.                                                           \\
+slice(x, l, h)           & Returns the substring of $x$ between $[l, h)$.                         \\
+store(n, v)              & Stores the value $v$ for the variable $n$.                             \\
+sub(x, y)                & Difference of $x$ and $y$.                                             \\
+write(k, v)              & Writes the value $v$ for the key $k$.                                  \\
+\end{tabular}
+\label{table:expressions}
+\end{table}
+
+\begin{table}[ht]
+\caption{Math Library}
+\centering
+\begin{tabular}{l | l}
+\hline\hline
+Function & Description                                                                            \\
+\hline
+abs(x)                   & Absolute value of $x$.                                                 \\
+acos(x)                  & Cosine inverse of $x$.                                                 \\
+acot(x)                  & Cotangent inverse of $x$.                                              \\
+acsc(x)                  & Consecant inverse of $x$.                                              \\
+asec(x)                  & Secant inverse of $x$.                                                 \\
+asin(x)                  & Sine inverse of $x$.                                                   \\
+atan(x)                  & Tangent inverse of $x$.                                                \\
+ceil(x)                  & Smallest integer greater than or equal to $x$.                         \\
+cos(x)                   & Cosine of $x$.                                                         \\
+cosh(x)                  & Hyperbolic cosine of $x$.                                              \\
+cot(x)                   & Cotangent of $x$.                                                      \\
+coth(x)                  & Hyperbolic cotangent of $x$.                                           \\
+csc(x)                   & Cosecant of $x$.                                                       \\
+csch(x)                  & Hyperbolic cosecant of $x$.                                            \\
+exp(x)                   & Exponential of $x$.                                                    \\
+floor(x)                 & Largest integer less than or equal to $x$.                             \\
+log(x)                   & Natural logarithm of $x$.                                              \\
+log(x, y)                & Log base $y$ of $x$.                                                   \\
+log10(x)                 & Log base 10 of $x$.                                                    \\
+log2(x)                  & Log base 2 of $x$.                                                     \\
+pow(x, y)                & Power of $x$ to the $y$.                                               \\
+random()                 & Uniformly random number on [0, 1).                                     \\
+round(x)                 & Closest integer to $x$.                                                \\
+round(x, y)              & Closest multiple of $y$ to $x$.                                        \\
+sec(x)                   & Secant of $x$.                                                         \\
+sech(x)                  & Hyperbolic secant of $x$.                                              \\
+signum(x)                & Sign of $x$.                                                           \\
+sin(x)                   & Sine of $x$.                                                           \\
+sinh(x)                  & Hyperbolic sine of $x$.                                                \\
+sqrt(x)                  & Square root of $x$.                                                    \\
+tan(x)                   & Tangent of $x$.                                                        \\
+\end{tabular}
+\label{table:math}
+\end{table}
+
+\printbibliography
 \end{document}

### 2018-04-06 caustic-library/src/main/scala/caustic/library/math/package.scala

@@ -42,6 +42,7 @@ package object math {
   def round (x: Value[Double], y: Value[Double]): Value[Double] = branch(x % y < y / 2, x - x % y, x - x % y + y)
   def sec   (x: Value[Double])                  : Value[Double] = One / cos(x)
   def sech  (x: Value[Double])                  : Value[Double] = One / cosh(x)
+  def signum(x: Value[Double])                  : Value[Int]    = branch(x === 0, 0, branch(x < 0, -1, 1))
   def sin   (x: Value[Double])                  : Value[Double] = runtime.sin(x)
   def sinh  (x: Value[Double])                  : Value[Double] = (exp(x) - exp(-x)) / Two
   def sqrt  (x: Value[Double])                  : Value[Double] = pow(x, Half)

### 2018-04-06 caustic-runtime/src/main/scala/caustic/runtime/Runtime.scala

@@ -32,6 +32,7 @@ class Runtime(database: Volume) {
     val buffer   = mutable.Map.empty[Key, Literal]
     val locals   = mutable.Map.empty[Key, Literal]
 
+    @tailrec
     def evaluate(iteration: Program): Try[Literal] = {
       // Fetch all keys that are read (for their value) and written (for their version), that have
       // not been read before (to avoid changes in value and version) to ensure that the evaluation
@@ -50,14 +51,16 @@ class Runtime(database: Volume) {
 
       // Fetch the keys, update the local snapshot, and reduce the program. If the result is a
       // literal then return, otherwise recurse on the partially evaluated program.
-      this.database.get(keys) flatMap { r =>
-        depends  ++= r.mapValues(r => r.version)
-        snapshot ++= r.mapValues(r => deserialize(r.value))
+      this.database.get(keys) match {
+        case Success(r) =>
+          depends  ++= r.mapValues(r => r.version)
+          snapshot ++= r.mapValues(r => deserialize(r.value))
 
-        reduce(List(iteration), List.empty) match {
-          case l: Literal => Success(l)
-          case o: Expression => evaluate(o)
-        }
+          reduce(List(iteration), List.empty) match {
+            case l: Literal => Success(l)
+            case o: Expression => evaluate(o)
+          }
+        case Failure(e) => Failure(e)
       }
     }
 

### 2018-04-07 caustic-assets/paper/main.tex
Updated Library Type System

@@
 \subsubsection{Math}
-The runtime natively supports just nine mathematical operations: $add$, $sub$, $mul$, $div$,
-$pow$, $log$, $floor$, $sin$, and $cos$. However, these primitive operations are sufficient to
-derive the entire Scala \href{https://www.scala-lang.org/api/2.12.1/scala/math/index.html}{math}
-library using various mathematical identities and Taylor series approximations. The $div$,
-$log$, $sin$, and $cos$ functions can actually be implemented in terms of the other primitive
-operations; however, native support for them was included in the runtime to improve performance.
-The standard library provides implementations for the functions enumerated in Table
-\ref{table:math}.
+The runtime natively supports just nine mathematical operations: $add$, $sub$, $mul$, $div$, $pow$,
+$log$, $floor$,$sin$, and $cos$. However, these primitive operations are sufficient to derive the
+entire Scala \href{https://www.scala-lang.org/api/2.12.1/scala/math/index.html}{math} library using
+various mathematical identities and Taylor series approximations. The $div$, $log$, $sin$, and $cos$
+functions can actually be implemented in terms of the other primitive operations; however, native
+support for them was included in the runtime to improve performance. The standard library provides
+implementations for the functions enumerated in Table \ref{table:math}.
 

### 2018-04-07 caustic-library/src/main/scala/caustic/library/collection/List.scala

@@ -2,6 +2,7 @@ package caustic.library.collection
 
 import caustic.library.control._
 import caustic.library.typing._
+import caustic.library.typing.Value._
 import caustic.runtime._
 
 import scala.language.reflectiveCalls
@@ -11,7 +12,7 @@ import scala.language.reflectiveCalls
  *
  * @param length Current size.
  */
-class List[T <: Primitive](length: Variable[Int]) {
+case class List[T <: Primitive](length: Variable[Int]) {
 
   /**
    * Returns the number of values in the list.
@@ -49,8 +50,12 @@ class List[T <: Primitive](length: Variable[Int]) {
    * @param context Parse context.
    */
   def foreach[U](f: (Value[Int], Value[T]) => U)(implicit context: Context): Unit = {
-    if (this.length.isInstanceOf[Remote[Int]]) context += prefetch(this.length.key, this.length)
-    val index = Local[Int](context.label())
+    // Prefetch all items if the list is remotely stored.
+    if (this.length.isInstanceOf[Variable.Remote[Int]])
+      context += prefetch(this.length.key, this.length, 1)
+
+    // Iterate through each item in the list.
+    val index = Variable.Local[Int](context.label())
     index := 0
 
     While (index < this.length) {
@@ -100,7 +105,7 @@ class List[T <: Primitive](length: Variable[Int]) {
    * @return Index of value in list or -1.
    */
   def indexOf(value: Value[T])(implicit context: Context): Value[Int] = {
-    val index = Local[Int](context.label())
+    val index = Variable.Local[Int](context.label())
     index := -1
     foreach { case (i, v) => If (index === -1 && v === value)(index := i) }
     index
@@ -114,7 +119,7 @@ class List[T <: Primitive](length: Variable[Int]) {
    * @return Whether or not the lists contain the same values.
    */
   def ===(that: List[T])(implicit context: Context): Value[Boolean] = {
-    val equal = Local[Boolean](context.label())
+    val equal = Variable.Local[Boolean](context.label())
     equal := true
     foreach { case (i, v) => equal := equal && this(i) === v }
     this.size === that.size && equal
@@ -137,18 +142,18 @@ class List[T <: Primitive](length: Variable[Int]) {
    * @return JSON representation.
    */
   def toJson(implicit context: Context): Value[String] = {
-    val json = Local[String](context.label())
+    val json = Variable.Local[String](context.label())
[…]

### 2018-04-07 caustic-library/src/main/scala/caustic/library/collection/Map.scala

@@ -2,6 +2,8 @@ package caustic.library.collection
 
 import caustic.library.control._
 import caustic.library.typing._
+import caustic.library.typing.Value._
+import caustic.runtime.prefetch
 
 import scala.language.reflectiveCalls
 
@@ -11,6 +13,12 @@ import scala.language.reflectiveCalls
  * @param keys Keys.
  */
 case class Map[A <: String, B <: Primitive](keys: Set[A]) {
+
+  /**
+   *
+   * @return
+   */
+  def length: Variable[Int] = this.keys.length
 
   /**
    * Returns the number of entries.
@@ -83,6 +91,11 @@ case class Map[A <: String, B <: Primitive](keys: Set[A]) {
    * @param context Parse context.
    */
   def foreach[U](f: (Value[A], Value[B]) => U)(implicit context: Context): Unit = {
+    // Prefetch all key-value pairs if the map is remotely stored.
+    if (this.length.isInstanceOf[Variable.Remote[Int]])
+      context += prefetch(this.length.key, this.size, 2)
+
+    // Iterate over each key-value pair in the map.
     this.keys.foreach(k => f(k, this(k)))
   }
 
@@ -104,7 +117,7 @@ case class Map[A <: String, B <: Primitive](keys: Set[A]) {
    * @return Whether or not the maps are equal.
    */
   def ===(that: Map[A, B])(implicit context: Context): Value[Boolean] = {
-    val equals = Local[Boolean](context.label())
+    val equals = Variable.Local[Boolean](context.label())
     foreach { case (k, v) => equals := equals && v === that(k) }
     equals && this.keys === that.keys
   }
@@ -116,18 +129,18 @@ case class Map[A <: String, B <: Primitive](keys: Set[A]) {
    * @return JSON representation.
    */
   def toJson(implicit context: Context): Value[String] = {
-    val json = Local[String](context.label())
+    val json = Variable.Local[String](context.label())
     json := "{"
 
     foreach { case (k, v) =>
       If (json === "{") {
-        json := json ++ k.quoted ++ ": " ++ v.toJson
+        json := json + k.quoted + ": " + v.toJson
       } Else {
-        json := json ++ ", " ++ k.quoted ++ ": " ++ v.toJson
+        json := json + ", " + k.quoted + ": " + v.toJson
       }
     }
 
-    json ++ "}"
+    json + "}"
   }
 
 }

### 2018-04-07 caustic-library/src/main/scala/caustic/library/collection/Set.scala

@@ -13,6 +13,12 @@ import scala.language.reflectiveCalls
 case class Set[T <: Primitive](toList: List[T]) {
 
   /**
+   *
+   * @return
+   */
+  def length: Variable[Int] = this.toList.length
+
+  /**
    * Returns the number of elements in the set.
    *
    * @return Number of elements.

### 2018-04-07 caustic-library/src/main/scala/caustic/library/control/Context.scala

@@ -27,7 +27,7 @@ case class Context(
    */
   def label(): Value[String] = {
     this.current += 1
-    string("$") ++ this.current
+    "$" + this.current
   }
 
 }

### 2018-04-07 caustic-library/src/main/scala/caustic/library/math/package.scala

@@ -1,52 +1,297 @@
 package caustic.library
 
 import caustic.library.typing._
+import caustic.library.typing.Value._
 import caustic.runtime
 import caustic.runtime._
 
 package object math {
 
-  // Scalars.
-  val Zero                                      : Value[Int]    = 0
-  val Half                                      : Value[Double] = 0.5
-  val One                                       : Value[Int]    = 1
-  val Two                                       : Value[Int]    = 2
-  val Ten                                       : Value[Int]    = 10
-  val Pi                                        : Value[Double] = scala.math.Pi
-  val E                                         : Value[Double] = scala.math.E
+  val E: Value[Double] = scala.math.E
+  val Pi: Value[Double] = scala.math.Pi
 
-  // Functions.
-  def abs   [X <: Double](x: Value[X])          : Value[X]      = branch(x < Zero, -x, x)
-  def acos  (x: Value[Double])                  : Value[Double] = Pi / 2 - asin(x)
-  def acot  (x: Value[Double])                  : Value[Double] = Pi / 2 - atan(x)
-  def acsc  (x: Value[Double])                  : Value[Double] = Pi / 2 - asec(x)
-  def asec  (x: Value[Double])                  : Value[Double] = acos(pow(x, -1))
-  def asin  (x: Value[Double])                  : Value[Double] = x + pow(x, 3) / 6 + pow(x, 5) * 3 / 40 + pow(x, 7) * 15 / 336
-  def atan  (x: Value[Double])                  : Value[Double] = asin(x / pow(x * x + 1, Half))
-  def ceil  (x: Value[Double])                  : Value[Int]    = floor(x) + One
-  def cos   (x: Value[Double])                  : Value[Double] = runtime.cos(x)
-  def cosh  (x: Value[Double])                  : Value[Double] = (exp(x) + exp(-x)) / Two
-  def cot   (x: Value[Double])                  : Value[Double] = cos(x) / sin(x)
-  def coth  (x: Value[Double])                  : Value[Double] = cosh(x) / sinh(x)
[…]

### 2018-04-07 caustic-library/src/main/scala/caustic/library/record/Field.scala

@@ -8,7 +8,7 @@ import caustic.library.typing._
 trait Field[Type] {
   type Container
   def apply(key: Variable[String], field: Value[String]): Container
-  def apply(ref: Reference[_], field: Value[String]): Container = apply(ref.pointer, field)
+  def apply[T](ref: Reference[T], field: Value[String]): Container = apply(ref.pointer, field)
 }
 
 object Field extends LowPriorityField {

### 2018-04-07 caustic-library/src/main/scala/caustic/library/record/Reference.scala

@@ -83,8 +83,7 @@ case class Reference[T](pointer: Variable[String]) {
     keys: Keys.Aux[Repr, KeysRepr],
     folder: LeftFolder.Aux[KeysRepr, ops.json.Args[T], ops.json.type, ops.json.Args[T]]
   ): Value[String] = {
-    val obj = string("{\"key\": ") ++ this.pointer.key.quoted
-    obj ++ keys().foldLeft(ops.json.Args(this, "", recursive))(ops.json).json ++ "}"
+    "{\"key\": " + this.pointer.key.quoted + keys().foldLeft(ops.json.Args(this, "", recursive))(ops.json).json + "}"
   }
 
   /**

### 2018-04-07 caustic-library/src/main/scala/caustic/library/record/ops/json.scala

@@ -30,7 +30,7 @@ object json extends Poly2 {
     selector: Selector.Aux[TRepr, FieldName, FieldType],
     field: Field.Aux[FieldType, Variable[FieldType]]
   ): Case.Aux[Args[T], FieldName, Args[T]] = at[Args[T], FieldName] { (x, f) =>
-    x.copy(json = x.json ++ ", \"" ++ f.name ++ "\": " ++ field(x.src, f.name).toJson)
+    x.copy(json = x.json + ", \"" + f.name + "\": " + field(x.src, f.name).toJson)
   }
 
   implicit def casePointer[
@@ -52,7 +52,7 @@ object json extends Poly2 {
     evidence: FieldType <:< Reference[FieldT]
   ): Case.Aux[Args[T], FieldName, Args[T]] = at[Args[T], FieldName] { (x, f) =>
     val json = if (x.recursive) field(x.src, f.name).toJson() else field(x.src, f.name).key
-    x.copy(json = x.json ++ ", \"" ++ f.name ++ "\": \"" ++ json ++ "\"")
+    x.copy(json = x.json + ", \"" + f.name + "\": \"" + json + "\"")
   }
 
   implicit def caseNested[
@@ -73,7 +73,7 @@ object json extends Poly2 {
     fieldFolder: LeftFolder.Aux[FieldKeys, Args[FieldT], json.type, Args[FieldT]],
     evidence: FieldType <:!< Reference[FieldT]
   ): Case.Aux[Args[T], FieldName, Args[T]] = at[Args[T], FieldName] { (x, f) =>
-    x.copy(json = x.json ++ ", \"" ++ f.name ++ "\": " ++ field(x.src, f.name).toJson())
+    x.copy(json = x.json + ", \"" + f.name + "\": " + field(x.src, f.name).toJson())
   }
 
 }

### 2018-04-07 caustic-library/src/main/scala/caustic/library/typing/Constant.scala

@@ -7,4 +7,4 @@ import caustic.runtime.Program
  *
  * @param get Program representation.
  */
-case class Constant[X <: Primitive](get: Program) extends Value[X]
+case class Constant[+T <: Primitive](get: Program) extends Value[T]

### 2018-04-07 caustic-library/src/main/scala/caustic/library/typing/Value.scala

@@ -1,6 +1,7 @@
 package caustic.library.typing
 
-import caustic.runtime.Program
+import caustic.runtime
+import caustic.runtime._
 
 /**
  * A scalar value.
@@ -13,5 +14,54 @@ trait Value[+T <: Primitive] {
    * @return Program representation.
    */
   def get: Program
+
+}
+
+object Value {
+
+  implicit class AdditionOps[X <: String](x: Value[X]) {
+    def +[Y <: X](y: Value[Y])(implicit evidence: Y <:< X): Value[X] = add(x, y)
+    def +[Y >: X <: Primitive](y: Value[Y]): Value[Y] = add(x, y)
+  }
+
+  implicit class ArithmeticOps[X >: Int <: Double](x: Value[X]) {
+    def unary_- : Value[X] = sub(0, x)
+    def -(y: Value[Int]): Value[X] = sub(x, y)
+    def -[Y >: X <: Double](y: Value[Y])(implicit evidence: X <:< Y): Value[Y] = sub(x, y)
+    def *(y: Value[Int]): Value[X] = mul(x, y)
+    def *[Y >: X <: Double](y: Value[Y])(implicit evidence: X <:< Y): Value[Y] = mul(x, y)
+    def /(y: Value[Int]): Value[X] = div(x, y)
+    def /[Y >: X <: Double](y: Value[Y])(implicit evidence: X <:< Y): Value[Y] = div(x, y)
+    def %(y: Value[Int]): Value[X] = mod(x, y)
+    def %[Y >: X <: Double](y: Value[Y])(implicit evidence: X <:< Y): Value[Y] = mod(x, y)
+    def toJson: Value[String] = branch(x, x, "null")
+  }
+
+  implicit class TextualOps[X >: String <: Primitive](x: Value[X]) {
+    def contains(y: Value[String]): Value[Boolean] = runtime.contains(x, y)
+    def indexOf(y: Value[String]): Value[Boolean] = runtime.indexOf(x, y)
+    def length: Value[Int] = runtime.length(x)
+    def matches(y: Value[String]): Value[Boolean] = runtime.matches(x, y)
+    def quoted: Value[String] = add("\"", add(x, "\""))
+    def substring(l: Value[Int], h: Value[Int] = length): Value[String] = slice(x, l, h)
+    def toJson: Value[String] = branch(x, x.quoted, "null")
+  }
+
[…]

### 2018-04-07 caustic-library/src/main/scala/caustic/library/typing/Variable.scala

@@ -1,12 +1,13 @@
 package caustic.library.typing
 
 import caustic.library.control.Context
+import caustic.library.typing.Value._
 import caustic.runtime._
 
 /**
  * A scalar variable.
  */
-trait Variable[T <: Primitive] extends Value[T] {
+sealed trait Variable[T <: Primitive] extends Value[T] {
 
   /**
    * Returns the name of the variable.
@@ -30,34 +31,45 @@ trait Variable[T <: Primitive] extends Value[T] {
    * @return Scoped variable.
    */
   def scope[U <: Primitive](x: Value[String]): Variable[U] = this match {
-    case Local(a) => Local[U](this.key ++ "/" ++ x)
-    case Remote(a) => Remote[U](this.key ++ "/" ++ x)
+    case Variable.Local(k) => Variable.Local(k + "/" + x)
+    case Variable.Remote(k) => Variable.Remote(k + "/" + x)
   }
 
 }
 
-/**
- * A local variable.
- *
- * @param key Variable name.
- */
-case class Local[T <: Primitive](key: Value[String]) extends Variable[T] {
+object Variable {
 
-  override def get: Program = load(key)
+  /**
+   * A local variable.
+   *
+   * @param key Variable name.
+   */
+  case class Local[T <: Primitive](key: Value[String]) extends Variable[T] {
+    override def get: Program = load(key)
+    override def set(that: Value[T])(implicit context: Context): Unit = context += store(key, that)
+  }
 
-  override def set(that: Value[T])(implicit context: Context): Unit = context += store(key, that)
+  /**
+   * A remote variable.
+   *
+   * @param key Variable name.
+   */
+  case class Remote[T <: Primitive](key: Value[String]) extends Variable[T] {
+    override def get: Program = read(key)
+    override def set(that: Value[T])(implicit context: Context): Unit = context += write(key, that)
+  }
 
-}
+  // Implicit Operations.
+  implicit class AssignmentOps[X <: Primitive](x: Variable[X]) {
+    def :=[Y <: X](y: Value[Y])(implicit context: Context): Unit = x.set(y)
+  }
 
-/**
- * A remote variable.
- *
- * @param key Variable name.
- */
-case class Remote[T <: Primitive](key: Value[String]) extends Variable[T] {
-
-  override def get: Program = read(key)
-
[…]

### 2018-04-07 caustic-library/src/main/scala/caustic/library/typing/package.scala

@@ -1,6 +1,6 @@
 package caustic.library
 
-import caustic.library.control.Context
+import caustic.library.typing.Value._
 import caustic.runtime
 import caustic.runtime._
 
@@ -8,76 +8,31 @@ import scala.language.implicitConversions
 
 package object typing {
 
-  val None                                          : Program           = runtime.Null
-  val True                                          : Value[Boolean]    = Constant(true)
-  val False                                         : Value[Boolean]    = Constant(false)
+  val None: Program = runtime.Null
+  val True: Value[Boolean] = runtime.True
+  val False: Value[Boolean] = runtime.False
 
   // Implicit Conversions.
-  implicit def program(x: Value[_])                 : Program           = x.get
-  implicit def value[X <: Primitive](x: Program)    : Value[X]          = Constant[X](x)
-  implicit def int(x: scala.Int)                    : Value[Int]        = Constant(runtime.real(x))
-  implicit def double(x: scala.Double)              : Value[Double]     = Constant(runtime.real(x))
-  implicit def boolean(x: scala.Boolean)            : Value[Boolean]    = Constant(runtime.flag(x))
-  implicit def string(x: java.lang.String)          : Value[String]     = Constant(runtime.text(x))
+  implicit def program[X <: Primitive](x: Value[X]): Program = x.get
+  implicit def value[X <: Primitive](x: Program): Value[X] = Constant[X](x)
+  implicit def int(x: scala.Int): Value[Int] = Constant(x)
+  implicit def int(x: scala.Long): Value[Int] = Constant(x)
+  implicit def double(x: scala.Double): Value[Double] = Constant(x)
+  implicit def string(x: java.lang.String): Value[String] = Constant(x)
+  implicit def boolean(x: scala.Boolean): Value[Boolean] = Constant(x)
 
   // Implicit Operations.
-  implicit class TextualOps[X <: String](x: Value[X]) {
-    def ++[Y <: X](y: Value[X]): Value[String] = add(x, y)
-    def quoted: Value[String] = add("\"", add(x, "\""))
[…]

### 2018-04-07 caustic-library/src/test/scala/caustic/library/math/PackageTest.scala

@@ -1,6 +1,7 @@
 package caustic.library.math
 
 import caustic.library.typing._
+import caustic.library.typing.Value._
 
 import org.junit.runner.RunWith
 import org.scalatest.junit.JUnitRunner

### 2018-04-07 caustic-library/src/test/scala/caustic/library/record/ReferenceTest.scala

@@ -20,7 +20,7 @@ class ReferenceTest extends FunSuite with Matchers {
   )
 
   test("References are statically typed") {
-    val foo = Reference[Foo](Local("foo"))
+    val foo = Reference[Foo](Variable.Local("foo"))
     "foo.get('a)" shouldNot compile
     "foo.get('x)" should compile
   }

### 2018-04-07 caustic-runtime/src/main/scala/caustic/runtime/Runtime.scala

@@ -1,6 +1,5 @@
 package caustic.runtime
 
-import caustic.runtime.Literal._
 import caustic.runtime.Runtime._
 
 import beaker.client._
@@ -54,7 +53,7 @@ class Runtime(database: Volume) {
       this.database.get(keys) match {
         case Success(r) =>
           depends  ++= r.mapValues(r => r.version)
-          snapshot ++= r.mapValues(r => deserialize(r.value))
+          snapshot ++= r.mapValues(r => Literal(r.value))
 
           reduce(List(iteration), List.empty) match {
             case l: Literal => Success(l)
@@ -88,8 +87,6 @@ class Runtime(database: Volume) {
         reduce(first :: Cons :: rest, second :: rem)
       case (Expression(Repeat, c :: b :: Nil) :: rest, rem) =>
         reduce(branch(c, cons(b, repeat(c, b)), Null) :: rest, rem)
-      case (Expression(Random, Nil) :: rest, rem) =>
-        reduce(rest, real(scala.util.Random.nextDouble()) :: rem)
       case ((e: Expression) :: rest, rem) =>
         reduce(e.operands.reverse ::: e.operator :: rest, rem)
 
@@ -102,12 +99,11 @@ class Runtime(database: Volume) {
       case (Store :: rest, k :: v :: rem) => reduce(rest, store(k, v) :: rem)
       case (Rollback :: _, (l: Literal) :: _) => throw Rollbacked(l)
       case (Rollback :: rest, x :: rem) => reduce(rest, rollback(x) :: rem)
-      case (Repeat :: rest, Flag(false) :: _ :: rem) => reduce(rest, rem)
+      case (Repeat :: rest, False :: _ :: rem) => reduce(rest, rem)
       case (Repeat :: rest, c :: b :: rem) => reduce(rest, repeat(c, b) :: rem)
-      case (Prefetch :: rest, k :: s :: rem) => reduce(rest, prefetch(k, s) :: rem)
+      case (Prefetch :: rest, k :: s :: r :: rem) => reduce(rest, prefetch(k, s, r) :: rem)
       case (Cons :: rest, f :: s :: rem) => reduce(rest, cons(f, s) :: rem)
       case (Branch :: rest, c :: p :: f :: rem) => reduce(rest, branch(c, p, f) :: rem)
-      case (Random :: rest, rem) => reduce(rest, random() :: rem)
 
       // Simplify String Expressions.
[…]

### 2018-04-07 caustic-runtime/src/main/scala/caustic/runtime/operator.scala

@@ -1,36 +1,35 @@
 package caustic.runtime
 
 /**
- * An enumeration of the various operators permitted in an [[Expression]].
+ * An enumeration of the various transformations permitted in an expressions.
  */
 sealed trait Operator
-case object Read      extends Operator  // Returns the value of key (1).            (Text) => Any
-case object Write     extends Operator  // Sets the value of key (1) to (2).        (Text, Any) => Null
-case object Load      extends Operator  // Returns the value of variable (1).       (Text) => Any
-case object Store     extends Operator  // Sets the value of variable (1) to (2).   (Text, Any) => Null
-case object Cons      extends Operator  // Evaluate (1) and then returns (2).       (Any, Any) => Any
-case object Repeat    extends Operator  // Evaluate (2) until (1) is not satisfied. (Flag, Any) => Null
-case object Prefetch  extends Operator  // Reads keys at (1)/i for i in [0, (2)).   (Text, Int) => Null
-case object Branch    extends Operator  // Evaluate (2) if (1) or (3) otherwise.    (Flag, Any, Any) => Any
-case object Rollback  extends Operator  // Discard changes and return (1).          (Any) => Any
-case object Random    extends Operator  // Returns a uniformly random on [0, 1).    () => Real
-case object Add       extends Operator  // Returns (1) plus (2).                    (Any, Any) => Any
-case object Sub       extends Operator  // Returns (1) minus (2).                   (Real, Real) => Real
-case object Mul       extends Operator  // Returns (1) times (2).                   (Real, Real) => Real
-case object Div       extends Operator  // Returns (1) divided by (2).              (Real, Real) => Real
-case object Mod       extends Operator  // Retu rns (1) mod (2).                     (Real, Real) => Real
-case object Pow       extends Operator  // Returns (1) raised to the (2).           (Real, Real) => Real
-case object Log       extends Operator  // Returns the natural logarithm of (1).    (Real) => Real
-case object Sin       extends Operator  // Returns the sine of (1) in radians.      (Real) => Real
-case object Cos       extends Operator  // Returns the cosine of (1) in radians.    (Real) => Real
-case object Floor     extends Operator  // Returns the floor of (1).                (Real) => Real
-case object Both      extends Operator  // Returns (1) && (2).                      (Flag, Flag) => Flag
-case object Either    extends Operator  // Returns (1) || (2).                      (Flag, Flag) => Flag
[…]

### 2018-04-07 caustic-runtime/src/main/scala/caustic/runtime/package.scala

@@ -6,243 +6,383 @@ import scala.language.implicitConversions
 
 package object runtime {
 
-  // Cache Literals.
-  val True = Flag(true)
-  val False = Flag(false)
-  val Empty = Text("")
+  val True: Flag = Flag(true)
+  val False: Flag = Flag(false)
+  val Empty: Text = Text("")
 
-  implicit def flag(value: Boolean): Literal = if (value) True else False
-  implicit def real(value: Int): Literal = Real(value)
-  implicit def real(value: Double): Literal = Real(value)
-  implicit def text(value: String): Literal = if (value.isEmpty) Empty else Text(value)
+  // Implicit Conversions.
+  implicit def flag(value: Boolean): Flag = if (value) True else False
+  implicit def real(value: Int): Real = Real(value)
+  implicit def real(value: Long): Real = Real(value)
+  implicit def real(value: Float): Real = Real(value)
+  implicit def real(value: Double): Real = Real(value)
+  implicit def text(value: String): Text = if (value.isEmpty) Empty else Text(value)
 
-  // Simplify Expressions.
-  def read(k: Program): Program = k match {
-    case Null => Null
-    case Real(a) => throw Fault(s"Read undefined for key $a")
-    case Flag(a) => throw Fault(s"Read undefined for key $a")
-    case _ => Expression(Read, k :: Nil)
+  /**
+   * Returns the sum of x and y.
+   *
+   * @param x Any.
+   * @param y Any.
+   * @return Any.
+   */
+  def add(x: Program, y: Program): Program = (x, y) match {
+    case (Null, Null) => Null
+    case (Text(a), b: Literal) => a + b.asString
+    case (a: Literal, Text(b)) => a.asString + b
+    case (a: Literal, b: Literal) => a.asDouble + b.asDouble
+    case _ => Expression(Add, x :: y :: Nil)
   }
 
-  def write(k: Program, v: Program): Program = (k, v) match {
-    case (Null, _) => throw Fault(s"Write undefined for key None")
-    case (Real(a), _) => throw Fault(s"Write undefined for key $a")
-    case (Flag(a), _) => throw Fault(s"Write undefined for key $a")
-    case _ => Expression(Write, k :: v :: Nil)
+  /**
[…]

### 2018-04-07 caustic-runtime/src/main/scala/caustic/runtime/program.scala

@@ -1,28 +1,29 @@
 package caustic.runtime
 
+import caustic.runtime.Runtime.Fault
 import java.io._
 import java.nio.charset.Charset
 import java.util.Base64
+import scala.util.Try
 
 /**
- * A transactional procedure. Programs are an abstract-syntax tree composed of [[Literal]] leaves
- * and [[Expression]] nodes.
+ * A procedure. Programs are abstract-syntax trees composed of literals and expressions.
  */
 sealed trait Program
 
 /**
- * A [[Literal]] transformation. Expressions map [[Literal]] operands to a [[Literal]] result
- * using their [[Operator]]. Because all expressions produce [[Literal]] results on [[Literal]]
- * arguments, every [[Program]] can eventually be reduced to a single [[Literal]].
+ * A literal transformation. Expressions map literal operands to a literal result using their
+ * operator. Because all expressions return literal values on literal arguments, every program can
+ * eventually be reduced to a single literal value.
  *
- * @param operator
- * @param operands
+ * @param operator Transformation.
+ * @param operands Arguments.
  */
 case class Expression(operator: Operator, operands: List[Program]) extends Program
 
 /**
- * A typed value. Literals contain primitive values of type [[Null]], [[Flag]], [[Real]], and
- * [[Text]] which correspond to null, boolean, double, and string respectively.
+ * A typed value. Literals contain primitive values of type null, flag, real, or text which
+ * correspond to null, boolean, double, and string respectively.
  */
 sealed trait Literal extends Program with Serializable
 case object Null extends Literal
@@ -36,26 +37,55 @@ object Literal {
   val charset: Charset = Charset.forName("UTF-8")
 
   /**
+   * Constructs a literal from the serialized base64 representation.
    *
-   * @param repr
-   * @return
+   * @param base64 Serialized representation.
+   * @return Literal.
    */
-  def deserialize(repr: String): Literal = {
-    val bytes  = new ByteArrayInputStream(Base64.getDecoder.decode(repr))
+  def apply(base64: String): Literal = {
+    val bytes = new ByteArrayInputStream(Base64.getDecoder.decode(base64))
     val stream = new ObjectInputStream(bytes)
     stream.readObject().asInstanceOf[Literal]
   }
 
-  /**
-   *
-   * @param literal
-   * @return
-   */
[…]

### 2018-04-07 caustic-runtime/src/test/scala/caustic/runtime/PackageTest.scala

@@ -10,60 +10,61 @@ import org.scalatest.{FunSuite, Matchers}
 class PackageTest extends FunSuite with Matchers {
 
   test("Literals are cached") {
-    flag(true) should be theSameInstanceAs True
-    flag(false) should be theSameInstanceAs False
+    True should be theSameInstanceAs True
+    False should be theSameInstanceAs False
     text("") should be theSameInstanceAs Empty
   }
 
   test("Expressions are simplified") {
     // Math Expressions.
-    add(real(6), real(9)) shouldEqual real(15)
-    add(text("a"), text("b")) shouldEqual text("ab")
-    add(flag(true), flag(false)) shouldEqual text("truefalse")
-    add(text("a"), real(0)) shouldEqual text("a0.0")
-    add(text("a"), flag(true)) shouldEqual text("atrue")
-    add(real(3.2), flag(true)) shouldEqual text("3.2true")
-    sub(real(9), real(6)) shouldEqual real(3)
-    mul(real(2), real(3)) shouldEqual real(6)
-    div(real(5), real(2)) shouldEqual real(2.5)
-    mod(real(5), real(2)) shouldEqual real(1)
-    pow(real(5), real(2)) shouldEqual real(25)
-    log(real(math.exp(2))) shouldEqual real(2)
-    sin(real(0.0)) shouldEqual real(0)
-    cos(real(0.0)) shouldEqual real(1)
-    floor(real(1.0)) shouldEqual real(1)
-    floor(real(1.5)) shouldEqual real(1)
-    floor(real(1.4)) shouldEqual real(1)
+    add(6, 9) shouldEqual real(15)
+    add("a", "b") shouldEqual text("ab")
+    add(true, false) shouldEqual real(1)
+    add("a", 0) shouldEqual text("a0")
+    add("a", true) shouldEqual text("atrue")
+    add(3.2, true) shouldEqual real(4.2)
+    sub(9, 6) shouldEqual real(3)
+    mul(2, 3) shouldEqual real(6)
+    div(5, 2) shouldEqual real(2.5)
+    mod(5, 2) shouldEqual real(1)
+    pow(5, 2) shouldEqual real(25)
+    log(math.exp(2)) shouldEqual real(2)
+    sin(0) shouldEqual real(0)
+    cos(0) shouldEqual real(1)
+    floor(1) shouldEqual real(1)
+    floor(1.5) shouldEqual real(1)
 
     // String Expressions.
-    runtime.length(text("Hello")) shouldEqual real(5.0)
[…]

### 2018-04-07 caustic-library/src/main/scala/caustic/library/record/Field.scala
Fixed Formatting

@@ -16,14 +16,15 @@ object Field extends LowPriorityField {
   // Scalar fields are contained in mutable variables.
   implicit def scalar[T <: Primitive]: Aux[T, Variable[T]] = new Field[T] {
     type Container = Variable[T]
-    override def apply(key: Variable[String], field: Value[String]): Variable[T] = key.scope(field)
+    override def apply(k: Variable[String], f: Value[String]): Variable[T] = k.scope(f)
   }
 
   // Pointers are dereferenced and contained in references.
   implicit def pointer[T]: Aux[Reference[T], Reference[T]] = new Field[Reference[T]] {
     override type Container = Reference[T]
-    override def apply(key: Variable[String], field: Value[String]): Reference[T] = Reference(key.scope(field))
+    override def apply(k: Variable[String], f: Value[String]): Reference[T] = Reference(k.scope(f))
   }
+
 }
 
 trait LowPriorityField {
@@ -33,7 +34,7 @@ trait LowPriorityField {
   // References are contained in references.
   implicit def nested[T]: Aux[T, Reference[T]] = new Field[T] {
     override type Container = Reference[T]
-    override def apply(key: Variable[String], field: Value[String]): Reference[T] = Reference(key.scope(field))
+    override def apply(k: Variable[String], f: Value[String]): Reference[T] = Reference(k.scope(f))
   }
 
 }

### 2018-04-07 caustic-library/src/main/scala/caustic/library/record/Reference.scala

@@ -83,7 +83,8 @@ case class Reference[T](pointer: Variable[String]) {
     keys: Keys.Aux[Repr, KeysRepr],
     folder: LeftFolder.Aux[KeysRepr, ops.json.Args[T], ops.json.type, ops.json.Args[T]]
   ): Value[String] = {
-    "{\"key\": " + this.pointer.key.quoted + keys().foldLeft(ops.json.Args(this, "", recursive))(ops.json).json + "}"
+    val json = "{\"key\": " + this.pointer.key.quoted
+    json + keys().foldLeft(ops.json.Args(this, "", recursive))(ops.json).json + "}"
   }
 
   /**

### 2018-04-07 caustic-library/src/main/scala/caustic/library/typing/Value.scala

@@ -19,6 +19,7 @@ trait Value[+T <: Primitive] {
 
 object Value {
 
+  // Implicit Operations.
   implicit class AdditionOps[X <: String](x: Value[X]) {
     def +[Y <: X](y: Value[Y])(implicit evidence: Y <:< X): Value[X] = add(x, y)
     def +[Y >: X <: Primitive](y: Value[Y]): Value[Y] = add(x, y)
